---
title: "Calculate Plasticity Metrics"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

# Setup

```{r setup}

library(colorRamp2)
library(ggplot2)
library(gridExtra)
library(grid)
library(lme4)

# load common garden data
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2024.Rdata')

str(dat)

# remove NA genotypes and genetic weirdos (possibly other species)
dat <- dat[! is.na(dat$Genotype),]
dat <- dat[! is.na(dat$k2_tricho),]

# print session info, including package versions
sessionInfo()

```

```{r markdown_settings}

# markdown settings
knitr::opts_chunk$set(fig.width = 12, fig.height = 8)

```

```{r setup_ggplot}

# ggplot settings
theme_set(theme_bw(base_size = 16))

```

# Functions to calculate plasticity

Testing several functions from Valladares et al 2006:

*RDPI*: Relative Distance Plasticity Index, averages pairwise differences for individuals of the same genotype in different environments

*RDPI median*: same as RDPI but pairwise genotype differences are summarized using the median instead of mean

*PPI*: Phenotypic Plasticity Index, (max - min)/max

```{r functions}

# input is a vector for trait measures of all individuals from one genotype (phenos), and same-length vector identifying the gardens those measures are from (gardens)

calc_rdpi <- function(phenos, gardens){
  
  df <- data.frame(phenos = phenos, gardens = gardens)
  
  # remove NAs
  df <- df[! is.na(df$phenos),]
  
  # for each row (tree), calculate relative distance values with all other trees except the one in the same garden (only across-environment comparisons)
  # only do this if there is >1 measure
  if(nrow(df) > 1){
    for(r in 1:nrow(df)){
    
    # rows to compare - exclude any row  in same garden
    compare <- df[df$gardens != df[r, 'gardens'],]
    
    # get phenotype value for this tree
    val1 <- df[r, 'phenos']
    
    # calculate values for each pair of phenotypes
    # (or relative distance)
    rel.dist <- vector()
    
    # loop through each pair for comparison
    for(n in 1:nrow(compare)){
      
      val2 <- compare[n, 'phenos']
      
      p.diff <- abs(val1-val2)
      p.sum <- val1+val2
      
      relDistN <- p.diff/p.sum
      
      #skip pairs that produce NaN values - happens when both heights are 0
      if(! is.nan(p.diff/p.sum)){
        rel.dist <- append(rel.dist, relDistN)
      }
    }
    }
    
  # calculate relative plasticity index by dividing by number of values
  return(sum(rel.dist)/length(rel.dist))
  } else { # if there is only one measure, return NA
    return(NA)
  }
  
}
  

# summarize with medians instead of means

calc_rdpi_med <- function(phenos, gardens){
  
  df <- data.frame(phenos = phenos, gardens = gardens)
  
  # remove NAs
  df <- df[! is.na(df$phenos),]
  
  # for each row (tree), calculate relative distance values with all other trees except the one in the same garden (only across-environment comparisons)
  # only do this if there is >1 measure
  if(nrow(df) > 1){
    for(r in 1:nrow(df)){
    
    # rows to compare - exclude any row  in same garden
    compare <- df[df$gardens != df[r, 'gardens'],]
    
    # get phenotype value for this tree
    val1 <- df[r, 'phenos']
    
    # calculate values for each pair of phenotypes
    # (or relative distance)
    rel.dist <- vector()
    
    # loop through each pair for comparison
    for(n in 1:nrow(compare)){
      
      val2 <- compare[n, 'phenos']
      
      p.diff <- abs(val1-val2)
      p.sum <- val1+val2
      
      relDistN <- p.diff/p.sum
      
      #skip pairs that produce NaN values - happens when both heights are 0
      if(! is.nan(p.diff/p.sum)){
        rel.dist <- append(rel.dist, relDistN)
      }
    }
    }
    
  # calculate relative plasticity index by dividing by number of values
  return(median(rel.dist))
  } else { # if there is only one measure, return NA
    return(NA)
  }
  
}

# phenoypic plasticity index
# instead of using genotype-garden means as in Valladares 2006, just use all values
# input is a vector for trait measures of all individuals from one genotype (phenos), garden IDs are not needed

calc_ppi <- function(phenos){
  
  phenos <- phenos[! is.na(phenos)]
  if(length(phenos) > 1){
    max <- max(phenos, na.rm = T)
    min <- min(phenos, na.rm = T)
    return(max-min)/max
  } else {
    return(NA)
  }
}

```

# Calculate plasticity by genotype

## Get genotype info, setup for saving plasticity to same dataframe

```{r get_genotype_info}

# make a dataframe with genotype info (will add plasticity metrics to this)
# I originally used these variables to plot their relationship with plasticity, but that is now in "plasticity_vs_ancestry_and_climate_correlations.Rmd" script

genos <- levels(dat$Genotype)
vars <- c("provenance_latitude", "provenance_longitude", "provenance_MAT", "provenance_MWMT", "provenance_MCMT", "provenance_TD", "provenance_MAP", "provenance_MSP", "provenance_AHM", "provenance_SHM", "provenance_DD_0", "provenance_DD5", "provenance_DD_18", "provenance_DD18", "provenance_NFFD", "provenance_bFFP", "provenance_eFFP", "provenance_FFP", "provenance_PAS", "provenance_EMT", "provenance_EXT", "provenance_MAR", "provenance_Eref", "provenance_CMD", "provenance_RH", "ID", "k2_balsam", "k2_tricho", "k3_balsam", "k3_coastal_tricho", "k3_interior_tricho", "genetic_PC1", "genetic_PC2", "genetic_PC3", "genetic_PC4", "genetic_PC5", "genetic_PC6","provenance_climate_PC1", "provenance_climate_PC2", "provenance_climate_PC3", "provenance_climate_PC4", "provenance_climate_PC5", "provenance_climate_PC6", "provenance_climate_PC7", "provenance_climate_PC8", "provenance_climate_PC9", "color_Pt", "color_k3", "interspecific_heterozygosity", "heterozygosity")

geno.info <- as.data.frame(matrix(nrow = length(genos), ncol = length(vars)))
rownames(geno.info) <- genos
colnames(geno.info) <- vars


# get genotype info
for(g in 1:length(genos)){
  
  geno <- genos[g]
  
  for(v in 1:length(vars)){
    
    var <- vars[v]
    
    # get first row for this genotype and extract variable
    geno.info[g, v] <- dat[match(geno, dat$Genotype), var]
    
  }
}

# add genotype
geno.info <- cbind.data.frame(genos, geno.info)
colnames(geno.info)[colnames(geno.info) == 'genos'] <- 'genotype'

head(geno.info)



```

## Transform trait data

If needed, transform trait data to meet assumptions of normality

```{r transform_trait}

# RDPI needs normally distributed data
# plot distribution of each trait

traits <- c("GrowthIncrement_2023", "DOY_Stage2_2023", "DOY_Stage3_2023", "DOY_last_budset_2023", "stage7_presence_2023", "growing_season_days_2023", "DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2023","DOY_Stage6_2024", "DOY_last_budset_2024", "stage7_presence_2024", "growing_season_days_2024", "licor_gsw", "licor_gbw","licor_PhiPS2", "licor_ETR", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023", "leaf_area_cm2_2023", "lower_stomata_count", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_count", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2", "lower_stomata_density_mm2", "stomata_density_both", "upper_stomata_density_over_total_density", "stomata_ratio")


par(mfrow = c(4,8))
for(n in 1:length(traits)){
  trait <- traits[n]
  hist(dat[,trait], main = trait, breaks = 'fd')
  
}

to_log <- c("DOY_Stage6_2023", "DOY_Stage6_2024", "leaf_mass_g_2023", "leaf_area_cm2_2023")

par(mfrow = c(1,2))
for(n in 1:length(to_log)){
  
  trait <- to_log[[n]]
  newcol <- paste(trait, 'log', sep = '_')
  dat[,newcol] <- log(dat[,trait])
  
  # plot comparison
  hist(dat[,trait], main = trait, breaks = 'fd')
  hist(dat[,newcol], main = 'log', breaks = 'fd')
  
}

# for traits where zero values are possible, need to add offset
to_log <- c("GrowthIncrement_2023",  "upper_stomata_count",  "upper_stomata_density_mm2",  "upper_stomata_density_over_total_density", "stomata_ratio")

par(mfrow = c(1,2))
for(n in 1:length(to_log)){
  
  trait <- to_log[[n]]
  newcol <- paste(trait, 'log', sep = '_')
  dat[,newcol] <- log(dat[,trait] + 0.01)
  
  # plot comparison
  hist(dat[,trait], main = trait, breaks = 'fd')
  hist(dat[,newcol], main = 'log', breaks = 'fd')
  
}

# new list of traits
traits <- c("GrowthIncrement_2023_log", "DOY_Stage2_2023", "DOY_Stage3_2023", "DOY_last_budset_2023", "stage7_presence_2023", "growing_season_days_2023", "DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2024", "DOY_last_budset_2024", "stage7_presence_2024", "growing_season_days_2024", "licor_gsw", "licor_gbw","licor_PhiPS2", "licor_ETR", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023_log", "leaf_area_cm2_2023_log", "lower_stomata_count", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_count_log", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2_log", "lower_stomata_density_mm2", "stomata_density_both", "upper_stomata_density_over_total_density_log", "stomata_ratio_log")

par(mfrow = c(4,8))
for(n in 1:length(traits)){
  trait <- traits[n]
  hist(dat[,trait], main = trait, breaks = 'fd')
  
}

# plot by garden

par(mfrow = c(3,4))
for(n in 1:length(traits)){
  
  trait <- traits[n]
  boxplot(dat[,trait] ~ factor(dat$MiniCG_Site_2023, exclude = c('OLLU', 'LOCK', 'NDSU', 'NWMO')),
          las = 2, main = trait, xlab = '', ylab = trait)
  
}

```

## Calculate Plasticity

Calculate RDPI, PPI, and summary metrics

```{r calc_plasticity_by_genotype}

# loop through genotypes and get info for each of them

# calculate plasticity for traits

traits <- c("GrowthIncrement_2023_log", "DOY_Stage2_2023", "DOY_Stage3_2023", "DOY_Stage6_2023", "DOY_last_budset_2023", "stage7_presence_2023", "growing_season_days_2023", "DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2024", "DOY_Stage7_2023", "DOY_Stage7_2024", "DOY_last_budset_2024", "stage7_presence_2024", "growing_season_days_2024", "licor_gsw", "licor_gbw","licor_PhiPS2", "licor_ETR", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023_log", "leaf_area_cm2_2023_log", "lower_stomata_count", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_count_log", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2_log", "lower_stomata_density_mm2", "stomata_density_both", "upper_stomata_density_over_total_density_log", "stomata_ratio_log")

plast <- list()

for(t in 1:length(traits)){
  
  pheno_colname <- traits[t]
  
  # get values for trait
  df <- as.data.frame(matrix(nrow = length(genos), ncol = 7))
  #rownames(df) <- paste('geno_', genos, sep = '') 
  rownames(df) <- genos
  colnames(df) <- c('rdpi', 'ppi', 'min', 'max', 'median', 'avg', 'sd')
  
  for(n in 1:length(genos)){
    
    geno <- genos[n]
    # get phenotype values
    sub <- dat[dat$Genotype == geno,]
    
    # get RDPI
    
    # todo
    # only calculate RDPI if trees exist in at least X sites
    tmp <- cbind(sub[,pheno_colname], as.character(sub$MiniCG_Site))
    if(length(unique(tmp[complete.cases(tmp), 2])) > 3){
      rdpi <- calc_rdpi(sub[,pheno_colname], sub$MiniCG_Site)
      df[n, 'rdpi'] <- rdpi
    }

    
    # get other info
    df[n, 'ppi'] <- calc_ppi(sub[,pheno_colname])
    df[n, 'min'] <- min(sub[,pheno_colname], na.rm = T)
    df[n, 'max'] <- max(sub[,pheno_colname], na.rm = T)
    df[n, 'median'] <- median(sub[,pheno_colname], na.rm = T)
    df[n, 'avg'] <- mean(sub[,pheno_colname], na.rm = T)
    df[n, 'sd'] <- sd(sub[,pheno_colname], na.rm = T)
    
    # replace Inf values in rdpi (caused by dividing by zero)
    df$rdpi[df$rdpi == Inf] <- NA
  }
  
  # save to list
  plast[[t]] <- df
  
  # also add rdpi and ppi as column 
  geno.info[,paste('plasticity_rdpi_', pheno_colname, sep = '')] <- df$rdpi
  geno.info[,paste('plasticity_ppi_', pheno_colname, sep = '')] <- df$ppi
  
  print(pheno_colname)
}

names(plast) <- traits

```

## Save

```{r save_plasticity}

write.csv(geno.info, row.names= F, file = 'data/clean/genotypePlasticityRDPI_2023-2024.csv')

# rename then save as R object
rdpi <- geno.info
save(rdpi, file = 'data/clean/genotypePlasticityRDPI_2023-2024.Rdata')

```


# Compare plasticity metrics

Check how different plasticity metrics are

RDPI and PPI are fairly similar for most traits.


```{r compare_plasticity, fig.width = 16, fig.height = 16}


plots <- list()

# does RDPI correlate with SD?
# mostly yes

for(n in 1:length(plast)){
  
  trait <- plast[[n]]
  # put into a df
  df <- data.frame(geno.info, trait)
  
  # calculate correlation
  cor <- cor.test(trait$rdpi, trait$sd)
  
  plots[[n]] <- ggplot(df, aes(x = rdpi, y = sd)) +
    geom_point(color = df$color_Pt, cex = 3, na.rm = TRUE) +
    ggtitle( paste(names(plast)[n], '\nR = ', round(cor$estimate, 2), '\np = ', round(cor$p.value, 3), sep = '')) +
    xlab('Plasticity (RDPI)') +
    ylab('SD')+
    #geom_smooth(method="loess", se=TRUE, color = 'black') +
    geom_smooth(method=lm, se=TRUE, na.rm = TRUE, color = 'black', lty = ifelse(cor$p.value <= 0.05, 1, 2)) +
    theme(plot.title=element_text(size=10))
  
  
}

#ggplot(plots)
grid.arrange(grobs = plots, ncol = 7)


# does RDPI correlate with PPI?
# mostly yes

plots <- list()

for(n in 1:length(plast)){
  
  trait <- plast[[n]]
  
  # put into a df
  df <- data.frame(geno.info, trait)
  
  # calculate correlation
  cor <- cor.test(trait$rdpi, trait$ppi)
  
  plots[[n]] <- ggplot(df, aes(x = rdpi, y = ppi)) +
    geom_point(color = df$color_Pt, cex = 3, na.rm = TRUE) +
    ggtitle( paste(names(plast)[n], '\nR = ', round(cor$estimate, 2), '\np = ', round(cor$p.value, 3), sep = '')) +
    xlab('Plasticity (RDPI)') +
    ylab('PPI')+
    geom_smooth(method=lm, se=TRUE,  na.rm = TRUE, color = 'black', lty = ifelse(cor$p.value <= 0.05 & ! is.na(cor$p.value), 1, 2)) +
    theme(plot.title=element_text(size=10))
  
}

#ggplot(plots)
grid.arrange(grobs = plots, ncol = 7)

```
