---
title: "Trait GxE Models"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

Tests for GxE effects on phenotypic traits. Unlike variance partitioning script, here we test the effects of specific climate and genetic variables.

Environment (E): garden MAT and MAP (for year of collection)

Genotype (G): Home MAT and MAP, genetic PCs 1-3

Genotype x Environment interaction (GxE): all possible combinations of G and E variables above

# Setup

```{r setup}

library(ggplot2)
library(lme4)
library(sjPlot) # model plots
library(circlize) # colorramp2
library(ComplexHeatmap)
library(effectsize) # standardize_parameters to standardize model coefficients
library(car) # Anova function
library(knitr) # kable() for tables


# load data
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2024.Rdata')


# set negative heights to NA
dat$GrowthIncrement_2024[dat$GrowthIncrement_2024 < 0] <- NA
dat$GrowthIncrement_2023[dat$GrowthIncrement_2023 < 0] <- NA
dat$GrowthIncrement_2022[dat$GrowthIncrement_2022 < 0] <- NA
dat$GrowthIncrement_2021[dat$GrowthIncrement_2021 < 0] <- NA

# remove NA genotypes and those without genetic info
dat <- dat[! is.na(dat$Genotype),]
dat <- dat[! is.na(dat$k2_tricho),]


# print session info, including package versions
sessionInfo()

```

```{r markdown_settings}

# markdown settings
knitr::opts_chunk$set(fig.width = 10, fig.height = 8)

```

```{r setup_ggplot}

# ggplot settings
theme_set(theme_bw(base_size = 16))

```

## Functions for plotting 

```{r plot_function}

# function to plot GxE interaction 
# plot genetics/climate vs trait, separate panel for each garden
plot_int <- function(trait, genetics, env, dat){
  
  dat$garden_sort <- reorder(dat$garden, dat[,env])
  
    p <- ggplot(data = dat, aes_string(x = genetics, y = trait, color = 'k2_tricho'))+
    geom_point(na.rm = TRUE) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    geom_smooth(method = lm, col = 'black', na.rm = TRUE)+
    facet_wrap(~garden_sort, drop = T) +
    ggtitle(paste(genetics, ' ~ ', trait,  '\n', 'gardens sorted by ', env, sep = ''))
  
  plot(p)
  
}


# plot variation in trait across gardens, sorting them by garden climate
# boxplot by garden
# use when the only variable is garden climate

plot_env <- function(trait, env, dat){
  
  dat$garden_sort <- reorder(dat$garden, dat[,env])
  #print(dat$garden_sort)
  
      p <- ggplot(data = dat, aes_string(x = 'garden_sort', y = trait))+
    geom_boxplot()+
    geom_jitter(na.rm = TRUE, aes(color = k2_tricho)) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    ggtitle(paste('garden ~ ', trait,  '\n', 'gardens sorted by ', env, sep = '')) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
    
  plot(p)
  
}


```


# Test for GxE effects on traits

Run each year separately, because we want to use the garden climate for the year of collection

## 2023 data

### Setup

```{r stats_setup_2023}

###############################################

# all vars measured in 2023
vars <- c("DOY_Stage2_2023", "Stage2_cGDD_2023", "DOY_Stage3_2023", "Stage3_cGDD_2023",  "DOY_Stage6_2023", "daylength_Stage6_2023", "DOY_Stage7_2023", "daylength_Stage7_2023", "DOY_last_budset_2023", "daylength_last_budset_2023", "stage7_presence_2023", "growing_season_days_2023", 
             "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", "leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", "upper_stomata_density_mm2", "stomata_ratio", "upper_stomata_presence", 
            "licor_gsw", "licor_gbw", "licor_ETR", "licor_Fs", "licor_Fm.", "licor_PhiPS2")

###############################################

# setup variables and put into a dataframe

# set climate and phenotype variables
garden_MAT <- dat$garden_MAT_2023
garden_MAT_2 <- dat$garden_MAT_2023^2
home_MAT <- dat$provenance_MAT
home_MAT_2 <- dat$provenance_MAT^2
garden_MAP <- dat$garden_MAP_2023
garden_MAP_2 <- dat$garden_MAP_2023^2
home_MAP <- dat$provenance_MAP
home_MAP_2 <- dat$provenance_MAP^2
home_lat <- dat$provenance_latitude
garden_lat <- dat$garden_Latitude

# random effects
block <- as.character(interaction(dat$MiniCG_Site, dat$block, drop = T))
#block <- as.character(dat$block)
Pt <- dat$Pt
k2_tricho <- dat$k2_tricho
genotype <- as.character(dat$Genotype)
garden <- as.character(dat$MiniCG_Site)
genetic_PC1 <- dat$genetic_PC1
genetic_PC2 <- dat$genetic_PC2
genetic_PC3 <- dat$genetic_PC3

# put in df
df <- data.frame(dat[,vars], garden_MAT, garden_MAT_2, home_MAT, home_MAT_2, garden_MAP, garden_MAP_2, home_MAP, home_MAP_2, home_lat, garden_lat, Pt, k2_tricho, genotype, garden, block, genetic_PC1, genetic_PC2, genetic_PC3)
rownames(df) <- dat$Unique_ID
str(df)


```


### Models

```{r model_with_all_vars_2023}

# run full model for all variables
vars <- c("DOY_Stage2_2023", "Stage2_cGDD_2023", "DOY_Stage3_2023", "Stage3_cGDD_2023",  "DOY_Stage6_2023", "DOY_Stage7_2023", "DOY_last_budset_2023", "stage7_presence_2023", "growing_season_days_2023", 
             "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", "leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", "upper_stomata_density_mm2", "stomata_ratio", "upper_stomata_presence", 
            "licor_gsw", "licor_gbw", "licor_ETR", "licor_Fs", "licor_Fm.", "licor_PhiPS2")



dput(vars)

# the binary variables won't converge
# Error: (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate

vars <- c("DOY_Stage2_2023", "Stage2_cGDD_2023", "DOY_Stage3_2023", "Stage3_cGDD_2023",  "DOY_Stage6_2023", "daylength_Stage6_2023", "DOY_Stage7_2023", "daylength_Stage7_2023", "DOY_last_budset_2023", "daylength_last_budset_2023",  "growing_season_days_2023", 
             "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", "leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", "upper_stomata_density_mm2", "stomata_ratio", 
            "licor_gsw", "licor_gbw", "licor_ETR", "licor_Fs", "licor_Fm.", "licor_PhiPS2")

# binary variables
vars_bin <- c( "stage7_presence_2023", "upper_stomata_presence")

mods.23 <- list()

for(n in 1:length(vars)){
  
  var <- vars[n]
  
  print(paste('Model for:', var))
  
  if(var %in% vars_bin){

    # these models failed to converge - don't run here
    # mod <- glm(paste0(var, '~ garden_MAT*home_MAT + garden_MAT*home_MAP +   garden_MAP*home_MAP + garden_MAP*home_MAT + genetic_PC1*garden_MAT + genetic_PC2*garden_MAT + genetic_PC3*garden_MAT +  genetic_PC1*garden_MAP + genetic_PC2*garden_MAP + genetic_PC3*garden_MAP + (1 | genotype) + (1 | garden/block) '),
    #             data = df,
    #             family = binomial()
    # )


  } else {
    #  full model with just linear interaction effects
    mod <- lmer(paste0(var, '~ garden_MAT*home_MAT + garden_MAT*home_MAP +   garden_MAP*home_MAP + garden_MAP*home_MAT + genetic_PC1*garden_MAT + genetic_PC2*garden_MAT + genetic_PC3*garden_MAT +  genetic_PC1*garden_MAP + genetic_PC2*garden_MAP + genetic_PC3*garden_MAP + (1 | genotype) + (1 | garden/block)'),
                data = df
    )
  }
    summary(mod)
    mod.info <- car::Anova(mod)
    
    print(mod.info)
    
    # save to list
    mods.23[[n]] <- mod
    
    # plot model effects
    p <- plot_model(mod, type = 'std', vline.color = "black", show.values = T, title = var)
    plot(p)
    
    # plots for significant factors
    pvals <- mod.info$`Pr(>Chisq)`
    if(min(pvals) <= 0.05){
      
      # get significant factors
      sig <- rownames(mod.info)[pvals <= 0.05]
      
      # plot for each significant factor
      for(f in 1:length(sig)){
        
        y <- var # y axis is always the trait
        
        # if it is an interaction effect, we need to include both variables in the plot
        if(grepl(':', sig[f])){
          
          # get two variables in interaction
          interaction_vars <- strsplit(sig[f], ':')[[1]]
          
          # env will start with 'garden' because it's always garden climate
          env <- grep('garden', interaction_vars, value = T)
          # the one without garden is the genetic effect, either PCs or home clim
          x <- grep('garden', interaction_vars, value = T, invert = T)
          
          p2 <- plot_int(trait = y, genetics = x, env = env, dat = df)
          #plot(p2)
          
          # if just a single variable, still plot gardens separately, but sort by MAT by default
        } else{
          
          # if single variable is garden climate, use that and plot boxplots
          if(grepl('garden', sig[f])){
            env <- sig[f]
            plot_env(trait = y, env = env, dat = df)
            
          # if single variable is a genotype characteristic, facet by garden sorted by MAT  
          } else {
            env <- 'garden_MAT'
          x <- sig[f]
           p2 <- plot_int(trait = y, genetics = x, env = env, dat = df)
          #plot(p2)
          }
        }
      }
    }
}

# plot all model effects together
# there are really too many variables for this to be useful
#plot_models(mods.23,std.est = T, m.labels = vars, p.shape = T, spacing = 1, vline.color = 'black', colors = cols25())

names(mods.23) <- vars

# setup dfs for effect sizes and p-values of each factor
# use last full model run in loop to get effect names - models in mods object have some effects reduced
rows <- names(fixef(mod))[-1] # names of fixed effects, minus intercept
# reorder row so garden clim is first
rows <- rows[c(1,4,2,3,5:17)]

colms <- vars

# effect size
df.eff.23 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.eff.23) <- colms
rownames(df.eff.23) <- rows
# standardized effect size
df.eff.std.23 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.eff.std.23) <- colms
rownames(df.eff.std.23) <- rows
# p values
df.pval.23 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.pval.23) <- colms
rownames(df.pval.23) <- rows
# p values
df.pval.all.23 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.pval.all.23) <- colms
rownames(df.pval.all.23) <- rows


# get p-values and effect sizes for each variable

for(n in 1:length(vars)){
  
  var <- vars[[n]]
  mod <- mods.23[[n]]
  # model info
  mod.info <- car::Anova(mod)
  # standardize coefficients
  eff.std <- effectsize::standardize_parameters(mod)
  
  # fixed effects, minus intercept
  #fes <- names(fixef(mod)[-1])
  # reorder so garden climates are first
  fes <- rownames(mod.info)
  
  # get their p-values and effect sizes for each fixed effect
    for(f in 1:length(fes)){
      
      fe <- fes[[f]]
      
      # check for significance
      pval <- mod.info[fe, 'Pr(>Chisq)']
      
      if(pval <= 0.05){
        df.eff.23[fe, var] <- fixef(mods.23[[var]])[fe]
        df.eff.std.23[fe,var] <- eff.std[eff.std$Parameter == fe, 'Std_Coefficient']
        df.pval.23[fe, var] <- mod.info[fe, 'Pr(>Chisq)']
      }
  }
}

# get all pvalues so they can be adjusted
for(n in 1:length(vars)){
  
  var <- vars[[n]]
  mod <- mods.23[[n]]
  #mod.info <- summary(mod)
  mod.info <- car::Anova(mod)

  # loop through retained variables and get their p-values and effect sizes
    for(f in 1:nrow(df.pval.all.23)){
      
      fe <- rownames(df.pval.all.23)[f]
      # get pval 
         df.pval.all.23[fe, var] <- mod.info[fe, 'Pr(>Chisq)']
         
  }
}

# adjust pvals
padj <- p.adjust(unlist(df.pval.all.23), method = 'BH')
df.pval.adj.23 <- as.data.frame(matrix(padj, ncol = ncol(df.pval.all.23)))
colnames(df.pval.adj.23) <- colnames(df.pval.all.23)
rownames(df.pval.adj.23) <- rownames(df.pval.all.23)

hist(unlist(df.pval.all.23))
hist(unlist(df.pval.adj.23))
plot(unlist(df.pval.all.23), unlist(df.pval.adj.23))

plot(unlist(df.eff.std.23), unlist(df.pval.all.23))

##################
# color scale based on all effect sizes
max_col <- max(unlist(df.eff.std.23), na.rm = T)
min_col <- min(unlist(df.eff.std.23), na.rm = T)

# make scale same at both ends
max_col <- max(abs(max_col), abs(min_col))
min_col <- min(abs(max_col), abs(min_col))*-1
max_ceil <- ceiling(max_col*10)/10 # take ceiling rounded to one decimal place - used for placing ticks on legend

colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))

# raw pvals
# cell_text <- df.pval
# cell_text[is.na(df.pval)] <- ''
# cell_text[df.pval > 0.05] <- ''
# cell_text[df.pval <= 0.05 & df.pval > 0.01] <- '*'
# cell_text[df.pval <= 0.01 & df.pval > 0.001] <- '**'
# cell_text[df.pval<= 0.001] <- '***'
# cell_text <- t(cell_text)

# adjusted pvalues
cell_text <- df.pval.adj.23
cell_text[is.na(df.pval.adj.23)] <- ''
cell_text[df.pval.adj.23 > 0.05] <- ''
cell_text[df.pval.adj.23 <= 0.05 & df.pval.adj.23 > 0.01] <- '*'
cell_text[df.pval.adj.23 <= 0.01 & df.pval.adj.23 > 0.001] <- '**'
cell_text[df.pval.adj.23<= 0.001] <- '***'
cell_text <- t(cell_text)


# heatmap

hm23 <- Heatmap(as.matrix(t(df.eff.std.23)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = T,
        column_title = 'Predictors',
        column_title_gp = gpar(fontface = 'bold'),
        row_title_gp = gpar(fontface = 'bold'),
        row_title = 'Trait',
        #left_annotation = ha_row,
        heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))}
        )


ComplexHeatmap::draw(hm23, padding = unit(c(2, 30, 2, 15), "mm"))



```

### Print model tables - 2023

```{r print_tables_2023}

kable(df.eff.23, label = 'Effects for 2023 traits')

kable(df.eff.std.23, label = 'Standardized effect sizes for 2023 traits')

kable(df.pval.all.23, label = 'p-values for 2023 traits')

kable(df.pval.adj.23, label = 'Adjusted p-values for 2023 traits')


```

### Test daylength effects on fall phenology

Daylength could be determining when genotypes set bud - check whether the daylength of budset is the same for a given genotype, or varies across gardens, which would suggest that other environmental factors contribute to budset timing 

```{r daylength}

# daylength of budset by genotype
# this varies across gardens, suggesting daylength is not the only factor controlling budset timing
ggplot(data = dat, aes(x = Genotype, y = daylength_Stage6_2023, color = k2_tricho))+
  geom_boxplot() +
    geom_jitter(width = 0.2, height = 0) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
  facet_wrap(~transect, scales = 'free_x')

ggplot(data = dat, aes(x = Genotype, y = daylength_Stage8_2023, color = k2_tricho))+
  geom_boxplot() +
    geom_jitter(width = 0.2, height = 0) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
  facet_wrap(~transect, scales = 'free_x')

# plot timing vs daylength by garden
ggplot(data = dat, aes(x = DOY_Stage6_2023, y = daylength_Stage6_2023, color = k2_tricho))+
    geom_vline(xintercept = 171, col = 'grey', lty = 2 )+  # day of solstice
    geom_point() +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    facet_wrap(~MiniCG_Site, drop = T)

ggplot(data = dat, aes(x = DOY_last_budset_2023, y = daylength_last_budset_2023, color = k2_tricho))+
    geom_vline(xintercept = 171, col = 'grey', lty = 2 )+  # day of solstice
    geom_point() +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    facet_wrap(~MiniCG_Site, drop = T)

ggplot(data = dat, aes(x = DOY_Stage6_2024, y = daylength_Stage6_2024, color = k2_tricho))+
  geom_vline(xintercept = 171, col = 'grey', lty = 2 )+  # day of solstice
    geom_point() +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    facet_wrap(~MiniCG_Site_2024, drop = T)

ggplot(data = dat, aes(x = DOY_last_budset_2024, y = daylength_last_budset_2024, color = k2_tricho))+
    geom_vline(xintercept = 171, col = 'grey', lty = 2 )+  # day of solstice
    geom_point() +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    facet_wrap(~MiniCG_Site_2024, drop = T)

```

## 2024 data

### Setup
```{r stats_setup_2024}

###############################################
# phenotypes
# all vars measured in 2024
vars <- c("DOY_Stage2_2024", "Stage2_cGDD_2024", "DOY_Stage3_2024", "Stage3_cGDD_2024", "DOY_Stage6_2024", "DOY_last_budset_2024",  "DOY_Stage7_2024", "stage7_presence_2024", "growing_season_days_2024")
###############################################

# setup variables and put into a dataframe

# set climate and phenotype variables
garden_MAT <- dat$garden_MAT_2024
garden_MAT_2 <- dat$garden_MAT_2024^2
home_MAT <- dat$provenance_MAT
home_MAT_2 <- dat$provenance_MAT^2
garden_MAP <- dat$garden_MAP_2024
garden_MAP_2 <- dat$garden_MAP_2024^2
home_MAP <- dat$provenance_MAP
home_MAP_2 <- dat$provenance_MAP^2
home_lat <- dat$provenance_latitude
garden_lat <- dat$garden_Latitude

# random effects
block <- as.character(interaction(dat$MiniCG_Site, dat$block, drop = T))
#block <- as.character(dat$block)
Pt <- dat$Pt
genotype <- as.character(dat$Genotype)
garden <- as.character(dat$MiniCG_Site)
genetic_PC1 <- dat$genetic_PC1
genetic_PC2 <- dat$genetic_PC2
genetic_PC3 <- dat$genetic_PC3

# put in df
df <- data.frame(dat[,vars], garden_MAT, garden_MAT_2, home_MAT, home_MAT_2, garden_MAP, garden_MAP_2, home_MAP, home_MAP_2, home_lat, garden_lat, Pt, genotype, garden, block, genetic_PC1, genetic_PC2, genetic_PC3)
rownames(df) <- dat$Unique_ID
str(df)


```


### Models

```{r model_with_all_vars_2024}

dput(vars)

# ignore stage 7 presence here because it's binary
vars <- c("DOY_Stage2_2024", "Stage2_cGDD_2024", "DOY_Stage3_2024", "Stage3_cGDD_2024", 
"DOY_Stage6_2024",  "DOY_Stage7_2024", "DOY_last_budset_2024",
"growing_season_days_2024")

mods.24 <- list()

for(n in 1:length(vars)){
  
  var <- vars[n]
  
  print(paste('Model for:', var))

  
  #  full model with just linear interaction effects
    mod <- lmer(paste0(var, '~ garden_MAT*home_MAT + garden_MAT*home_MAP +   garden_MAP*home_MAP + garden_MAP*home_MAT + genetic_PC1*garden_MAT + genetic_PC2*garden_MAT + genetic_PC3*garden_MAT +  genetic_PC1*garden_MAP + genetic_PC2*garden_MAP + genetic_PC3*garden_MAP + (1 | genotype) + (1 | garden/block)'),
                        data = df
  )
  
  mod.info <- Anova(mod)
  
  print(mod.info)

    
  # save to list
  mods.24[[n]] <- mod
  
  p <- plot_model(mod, type = 'std', vline.color = "black", show.values = T, title = var)
  plot(p)
  
  
  # plots for significant factors
    pvals <- mod.info$`Pr(>Chisq)`
    if(min(pvals) <= 0.05){
      
      # get significant factors
      sig <- rownames(mod.info)[pvals <= 0.05]
      
      # plot for each significant factor
      for(f in 1:length(sig)){
        
        y <- var # y axis is always the trait
        
        # if it is an interaction effect, we need to include both variables in the plot
        if(grepl(':', sig[f])){
          
          # get two variables in interaction
          interaction_vars <- strsplit(sig[f], ':')[[1]]
          
          # env will start with 'garden' because it's always garden climate
          env <- grep('garden', interaction_vars, value = T)
          # the one without garden is the genetic effect, either PCs or home clim
          x <- grep('garden', interaction_vars, value = T, invert = T)
          
          p2 <- plot_int(trait = y, genetics = x, env = env, dat = df)
          #plot(p2)
          
          # if just a single variable, still plot gardens separately, but sort by MAT by default
        } else{
          
          # if single variable is garden climate, use that and plot boxplots
          if(grepl('garden', sig[f])){
            env <- sig[f]
            plot_env(trait = y, env = env, dat = df)
            
          # if single variable is a genotype characteristic, facet by garden sorted by MAT  
          } else {
            env <- 'garden_MAT'
          x <- sig[f]
           p2 <- plot_int(trait = y, genetics = x, env = env, dat = df)
          #plot(p2)
          }
        }
      }
    }
  
}


#plot_models(mods.24,std.est = T, m.labels = vars, p.shape = T, spacing = 1, vline.color = 'black', colors = cols25())

names(mods.24) <- vars

# setup dfs for effect sizes and p-values of each factor
# use last full model run in loop to get effect names - models in mods object have some effects reduced
rows <- names(fixef(mod))[-1] # names of fixed effects, minus intercept
# reorder row so garden clim is first
rows <- rows[c(1,4,2,3,5:17)]
colms <- vars

# effect size
df.eff.24 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.eff.24) <- colms
rownames(df.eff.24) <- rows
# standardized effect size
df.eff.std.24 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.eff.std.24) <- colms
rownames(df.eff.std.24) <- rows
# p values
df.pval.24 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.pval.24) <- colms
rownames(df.pval.24) <- rows
# all p values
df.pval.all.24 <- as.data.frame(matrix(ncol = length(colms), nrow = length(rows)))
colnames(df.pval.all.24) <- colms
rownames(df.pval.all.24) <- rows

# get variables and their p-values 

for(n in 1:length(vars)){
  
  var <- vars[[n]]
  mod <- mods.24[[n]]
  mod.info <- car::Anova(mod)
  # standardize coefficients
  eff.std <- effectsize::standardize_parameters(mod)
  
  # fixed effects, minus intercept
  fes <- names(fixef(mod)[-1])
  

  
  fes <- rownames(mod.info)
  
  # get their p-values and effect sizes for each fixed effect
    for(f in 1:length(fes)){
      
      fe <- fes[[f]]
      
      # check for significance
      pval <- mod.info[fe, 'Pr(>Chisq)']
      
      if(pval <= 0.05){
        df.eff.24[fe, var] <- fixef(mods.24[[var]])[fe]
        df.eff.std.24[fe,var] <- eff.std[eff.std$Parameter == fe, 'Std_Coefficient']
        df.pval.24[fe, var] <- mod.info[fe, 'Pr(>Chisq)']
      }
  }
}

# get all pvalues so they can be adjusted
for(n in 1:length(vars)){
  
  var <- vars[[n]]
  mod <- mods.24[[n]]
  mod.info <- car::Anova(mod)

  # loop through retained variables and get their p-values and effect sizes
    for(f in 1:nrow(df.pval.all.24)){
      
      fe <- rownames(df.pval.all.24)[f]
      # get pval 
         df.pval.all.24[fe, var] <- mod.info[fe, 'Pr(>Chisq)']
         
        # for car
         
  }
}

# adjust pvals
padj <- p.adjust(unlist(df.pval.all.24), method = 'BH')
df.pval.adj.24 <- as.data.frame(matrix(padj, ncol = ncol(df.pval.all.24)))
colnames(df.pval.adj.24) <- colnames(df.pval.all.24)
rownames(df.pval.adj.24) <- rownames(df.pval.all.24)

hist(unlist(df.pval.all.24))
hist(unlist(df.pval.adj.24))
plot(unlist(df.pval.all.24), unlist(df.pval.adj.24))

plot(unlist(df.eff.std.24), unlist(df.pval.all.24))

##################
# color scale based on all effect sizes
max_col <- max(unlist(df.eff.std.24), na.rm = T)
min_col <- min(unlist(df.eff.std.24), na.rm = T)

# make scale same at both ends
max_col <- max(abs(max_col), abs(min_col))
min_col <- min(abs(max_col), abs(min_col))*-1
max_ceil <- ceiling(max_col*10)/10 # take ceiling rounded to one decimal place - used for placing ticks on legend

colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))

# raw pvals
# cell_text <- df.pval
# cell_text[is.na(df.pval)] <- ''
# cell_text[df.pval > 0.05] <- ''
# cell_text[df.pval <= 0.05 & df.pval > 0.01] <- '*'
# cell_text[df.pval <= 0.01 & df.pval > 0.001] <- '**'
# cell_text[df.pval<= 0.001] <- '***'
# cell_text <- t(cell_text)

# adjusted pvalues
cell_text <- df.pval.adj.24
cell_text[is.na(df.pval.adj.24)] <- ''
cell_text[df.pval.adj.24 > 0.05] <- ''
cell_text[df.pval.adj.24 <= 0.05 & df.pval.adj.24 > 0.01] <- '*'
cell_text[df.pval.adj.24 <= 0.01 & df.pval.adj.24 > 0.001] <- '**'
cell_text[df.pval.adj.24<= 0.001] <- '***'
cell_text <- t(cell_text)


# heatmap for 2024 traits
hm24 <- Heatmap(as.matrix(t(df.eff.std.24)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = T,
        column_title = 'Predictors',
        column_title_gp = gpar(fontface = 'bold'),
        row_title_gp = gpar(fontface = 'bold'),
        row_title = 'Trait',
        #left_annotation = ha_row,
        heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))}
        )


ComplexHeatmap::draw(hm24, padding = unit(c(2, 30, 2, 15), "mm"))


```

### Print model tables - 2024

```{r print_tables_2024}

kable(df.eff.24, label = 'Effects for 2024 traits')

kable(df.eff.std.24, label = 'Standardized effect sizes for 2024 traits')

kable(df.pval.all.24, label = 'p-values for 2024 traits')

kable(df.pval.adj.24, label = 'Adjusted p-values for 2024 traits')


```

# Summary heatmap

```{r plot_years_together}

# check that effects match
cbind(rownames(df.eff.std.23), rownames(df.eff.std.24))

# merge effect sizes
df.eff.both <- cbind.data.frame(df.eff.std.23, df.eff.std.24)
head(df.eff.both)

# merge pvalues
df.pval.adj.both <- cbind.data.frame(df.pval.adj.23, df.pval.adj.24)

rownames(df.pval.adj.both) <- rownames(df.eff.both)
colnames(df.pval.adj.both) <- colnames(df.eff.both)

df.pval.both <- cbind.data.frame(df.pval.23, df.pval.24)
rownames(df.pval.both) <- rownames(df.eff.both)
colnames(df.pval.both) <- colnames(df.eff.both)

# reorder rows
dput(colnames(df.eff.both))
ord <- c("growing_season_days_2023", "growing_season_days_2024", "DOY_Stage2_2023", "Stage2_cGDD_2023", "DOY_Stage2_2024", "Stage2_cGDD_2024", "DOY_Stage3_2023", "Stage3_cGDD_2023",  "DOY_Stage3_2024", "Stage3_cGDD_2024","DOY_Stage6_2023", "DOY_Stage6_2024", "DOY_Stage7_2023", "DOY_Stage7_2024", "DOY_last_budset_2023", "DOY_last_budset_2024", 
 "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", 
"leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", 
"upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", 
"upper_stomata_density_mm2",  "licor_gsw", "stomata_ratio", 
"licor_ETR", "licor_Fm.", "licor_Fs", "licor_gbw",  
  "licor_PhiPS2"  
)

df.eff.both <- df.eff.both[,ord]
df.pval.adj.both <- df.pval.adj.both[,ord]


# clean up trait names

trait_names <- colnames(df.eff.both)
trait_names <- gsub('Stage2_cGDD', 'Stage2_Bud_Flush_cGDD', trait_names)
trait_names <- gsub('DOY_Stage2', 'Stage2_Bud_Flush_DOY', trait_names)
trait_names <- gsub('DOY_Stage3', 'Stage3_Leaf_Emergence_DOY', trait_names)
trait_names <- gsub('DOY_Stage6', 'Stage6_First_Budset_DOY', trait_names)
trait_names <- gsub('DOY_Stage7', 'Stage7_Lammas_Growth_DOY', trait_names)
#trait_names <- gsub('stage7_presence', 'Stage7_Lammas_Growth_presence', trait_names)
trait_names <- gsub('DOY_Stage8', 'Stage8_Budset_after_Lammas_DOY', trait_names)
trait_names <- gsub('DOY_last_budset', 'Stage8_Final_Budset_DOY', trait_names)
trait_names <- gsub('growing_season_days', 'Growing_Season_Days', trait_names)

trait_names <- gsub('licor_', '', trait_names)
trait_names <- gsub('Fm.', "Fm'", trait_names)
trait_names <- gsub('upper', "Adaxial", trait_names)
trait_names <- gsub('lower', "Abaxial", trait_names)
trait_names <- gsub('stomata_ratio', "Stomata_ratio", trait_names)
trait_names <- gsub('leaf', 'Leaf', trait_names)

# remove units
trait_names <- gsub('_cm2', '', trait_names)
trait_names <- gsub('_cm', '', trait_names)
trait_names <- gsub('_mm2', '', trait_names)
trait_names <- gsub('_m2', '', trait_names)
trait_names <- gsub('_mm', '', trait_names)
trait_names <- gsub('_um', '', trait_names)
trait_names <- gsub('_g', '', trait_names)

trait_names <- gsub('_', ' ', trait_names)

# check
cbind(colnames(df.eff.both), trait_names)


# rename effects
effect_names <- rownames(df.eff.both)
effect_names <- gsub('home', 'Home', effect_names)
effect_names <- gsub('garden', 'Garden', effect_names)
effect_names <- gsub('genetic', 'Genetic', effect_names)
effect_names <- gsub(':', ' x ', effect_names)
effect_names <- gsub('_', ' ', effect_names)
# check
cbind(rownames(df.eff.both), effect_names)

# adjusted pvalues
cell_text <- df.pval.adj.both
cell_text[is.na(df.pval.adj.both)] <- ''
cell_text[df.pval.adj.both > 0.05] <- ''
cell_text[df.pval.adj.both <= 0.05 & df.pval.adj.both > 0.01] <- '*'
cell_text[df.pval.adj.both <= 0.01 & df.pval.adj.both > 0.001] <- '**'
cell_text[df.pval.adj.both<= 0.001] <- '***'
cell_text <- t(cell_text)

# color scale based on all effect sizes
max_col <- max(unlist(df.eff.both), na.rm = T)
min_col <- min(unlist(df.eff.both), na.rm = T)

# make scale same at both ends
max_col <- max(abs(max_col), abs(min_col))
min_col <- min(abs(max_col), abs(min_col))*-1
max_ceil <- ceiling(max_col*10)/10 # take ceiling rounded to one decimal place - used for placing ticks on legend

colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))

# column annotation - GxE categories
col_cat <-  c(rep('E', 2), rep('G', 5), rep('GxE', 10))
ha_col <- columnAnnotation(Effect_Category = col_cat,
                        show_annotation_name = F,
                            col = list(Effect_Category = c('E' = "#015b58", 'G' = "#4760b8", 'GxE' = "#380069" )))


# row annotation - trait categories
cats <- c(rep('Phenology', 16),  rep('Leaf Morphology', 4), rep('Stomata', 6), rep('Photosynthesis', 5))
cbind(colnames(df.eff.both), cats)

# choose colors
ha_row <- rowAnnotation(Trait_Category = cats,
                        show_annotation_name = F,
                            col = list(Trait_Category = c('Phenology' = "#e2e260", 'Photosynthesis' = "#2d4030", 'Leaf Morphology' = "#516823", 'Stomata' =  "#88a2b9")))


# combined heatmap
hm_both <- Heatmap(as.matrix(t(df.eff.both)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        row_labels = trait_names,
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = F,
        column_title = 'Trait Predictors',
        column_title_gp = gpar(fontface = 'bold'),
        column_labels = effect_names,
        row_title_gp = gpar(fontface = 'bold'),
        row_title = 'Trait',
        left_annotation = ha_row,
        top_annotation = ha_col,
        heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))}
        )


#png(file = 'results/plasticity/GxE_full_models_heatmap_2023-2024.png', height = 10, width = 12, res = 300, units = 'in')
#pdf(file = 'results/plasticity/GxE_full_models_heatmap_2023-2024.pdf', height = 10, width = 12)
ComplexHeatmap::draw(hm_both, padding = unit(c(2, 30, 2, 15), "mm"))
#dev.off()


# save heatmap object and matrix used for it
# can use to combine with plasticity heatmap
gxe_hm <- list(effect = df.eff.both,
               pval = df.pval.both,
               pval_adj = df.pval.adj.both,
               hm_text = cell_text,
               row_annot = ha_row,
               col_annot = ha_col,
               heatmap = hm_both)


#save(gxe_hm, file = 'results/plasticity/GxE_traits_heatmap_data.Rdata')

```

