---
title: "Plasticity variation by genotype"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

# Setup

```{r setup}

library(ggplot2)
library(gridExtra) # for plot panels
library(cowplot) # for extracting legend from plots
library(viridis) # colors
library(circlize) # color palette
library(bbmle) # AICtab
library(ComplexHeatmap) # heatmaps
library(lm.beta) # standardize model coefficients
library(psych) # corPlot()
library(knitr) # tables with kable()


# load plasticity data
load('data/clean/genotypePlasticityRDPI_2023-2024.Rdata')
dat <- rdpi
rm(rdpi)

# remove genotypes without genetic info
dat <- dat[! is.na(dat$k2_tricho),]
str(dat)

# print session info, including package versions
sessionInfo()

```

```{r markdown_settings}

# markdown settings
knitr::opts_chunk$set(fig.width = 10, fig.height = 8)

```

```{r setup_ggplot}

# ggplot settings
theme_set(theme_bw(base_size = 16))

```

# Preliminary correlation matrix

```{r preliminary_plots, fig.width = 12, fig.height = 12}

vars <- c("plasticity_rdpi_GrowthIncrement_2023_log", "plasticity_rdpi_DOY_Stage2_2023", "plasticity_rdpi_DOY_Stage3_2023", "plasticity_rdpi_DOY_last_budset_2023", "plasticity_rdpi_stage7_presence_2023", "plasticity_rdpi_growing_season_days_2023", "plasticity_rdpi_DOY_Stage2_2024", "plasticity_rdpi_DOY_Stage3_2024", "plasticity_rdpi_DOY_Stage6_2024", "plasticity_rdpi_DOY_last_budset_2024", "plasticity_rdpi_growing_season_days_2024", "plasticity_rdpi_licor_gsw", "plasticity_rdpi_licor_gbw", "plasticity_rdpi_licor_PhiPS2", "plasticity_rdpi_licor_ETR", "plasticity_rdpi_licor_Fs", "plasticity_rdpi_licor_Fm.", "plasticity_rdpi_leaf_thickness_avg_mm_2023", "plasticity_rdpi_LMA_g_m2_2023", "provenance_MAT",  "provenance_MAP", "provenance_TD", "provenance_latitude", "genetic_PC1", "genetic_PC2", "genetic_PC3")

# correlation matrix
cors <- cor(dat[,vars], use = 'pairwise.complete.obs')

# heatmap of correlations
rows <- c("plasticity_rdpi_GrowthIncrement_2023_log", "plasticity_rdpi_DOY_Stage2_2023", "plasticity_rdpi_DOY_Stage3_2023", "plasticity_rdpi_DOY_last_budset_2023", "plasticity_rdpi_stage7_presence_2023", "plasticity_rdpi_growing_season_days_2023", "plasticity_rdpi_DOY_Stage2_2024", "plasticity_rdpi_DOY_Stage3_2024", "plasticity_rdpi_DOY_Stage6_2024", "plasticity_rdpi_DOY_last_budset_2024", "plasticity_rdpi_growing_season_days_2024", "plasticity_rdpi_licor_gsw", "plasticity_rdpi_licor_gbw", "plasticity_rdpi_licor_PhiPS2", "plasticity_rdpi_licor_ETR", "plasticity_rdpi_licor_Fs", "plasticity_rdpi_licor_Fm.", "plasticity_rdpi_leaf_thickness_avg_mm_2023", "plasticity_rdpi_LMA_g_m2_2023")
columns <- c("provenance_MAT",  "provenance_MAP", "provenance_TD", "provenance_latitude", "genetic_PC1", "genetic_PC2", "genetic_PC3")

# trait plasticity vs predictors
corPlot(cors[rows, columns], MAR = 16, las = 2, symmetric = FALSE)

# correlations of predictors
corPlot(cors[columns, columns], MAR = 16, las = 2)

# correlations of trait plasticity
corPlot(cors[rows, rows], MAR = 16, las = 2)

```


# Relationship between plasticity and home climate / species ancestry

## Statistical models 

Test whether home climate (MAP, MAT, and TD) and genetic structure (PCs 1-3) predict trait plasticity using a linear regression

```{r calc_regressions}

# plasticity traits
traits <- c("plasticity_rdpi_growing_season_days_2023", "plasticity_rdpi_growing_season_days_2024", "plasticity_rdpi_DOY_Stage2_2023", "plasticity_rdpi_DOY_Stage2_2024","plasticity_rdpi_DOY_Stage3_2023", "plasticity_rdpi_DOY_Stage3_2024", "plasticity_rdpi_DOY_Stage6_2023",  "plasticity_rdpi_DOY_Stage6_2024", "plasticity_rdpi_DOY_Stage7_2023",  "plasticity_rdpi_DOY_Stage7_2024", "plasticity_rdpi_stage7_presence_2023", "plasticity_rdpi_stage7_presence_2024", "plasticity_rdpi_DOY_last_budset_2023", "plasticity_rdpi_DOY_last_budset_2024",  
 "plasticity_rdpi_leaf_thickness_avg_mm_2023", "plasticity_rdpi_leaf_area_cm2_2023_log","plasticity_rdpi_leaf_mass_g_2023_log", "plasticity_rdpi_LMA_g_m2_2023", "plasticity_rdpi_lower_stomata_pore_length_mean_um", "plasticity_rdpi_upper_stomata_pore_length_mean_um", "plasticity_rdpi_lower_stomata_density_mm2",
 "plasticity_rdpi_upper_stomata_density_mm2_log", "plasticity_rdpi_upper_stomata_presence",   "plasticity_rdpi_upper_stomata_density_over_total_density_log",
 "plasticity_rdpi_licor_gsw",
  "plasticity_rdpi_licor_ETR","plasticity_rdpi_licor_Fm.",  "plasticity_rdpi_licor_Fs", "plasticity_rdpi_licor_gbw",  "plasticity_rdpi_licor_PhiPS2")


# possible predictors of plasticity
preds <- c('genetic_PC1', 'genetic_PC2', 'genetic_PC3', 'provenance_MAT', 'provenance_MAP', 'provenance_TD')

# set up dataframes for saving model info (effects and pvals) within a list
res.lr <- list() # results for linear regressions
# setup template dataframe
df <- as.data.frame(matrix(ncol = length(traits), nrow = length(preds)))
colnames(df) <- traits
rownames(df) <- preds

# add duplicate dataframes to list for saving effect size and pvalues
res.lr$eff <- df
res.lr$eff.sig <- df
res.lr$eff.std <- df
res.lr$eff.std.sig <- df
res.lr$pvals <- df
res.lr$rsq <- df
res.lr$cors <- df
res.lr$cors.sig <- df

# loop through each trait/predictor combination, run a regression, and extract info from model

for(t in 1:length(traits)){
  
  trait <- traits[t]
  
  for(p in 1:length(preds)){
    
    pred <- preds[p]
    
    # model
    form <- paste0(trait, '~', pred)
    mod <- lm(form, data = dat)
    mod.info <- summary(mod)
    print(paste('Model:', trait, '~', pred))
    print(mod.info)
    
    res.lr$rsq[pred, trait] <- mod.info$adj.r.squared
    res.lr$pvals[pred, trait] <- mod.info$coefficients[pred, 'Pr(>|t|)']
    res.lr$eff[pred, trait] <- mod.info$coefficients[pred, 'Estimate']
    
    # standardize coefficients
    # coef for predictor is second object after intercept
    res.lr$eff.std[pred,trait] <- lm.beta(mod)$standardized.coefficients[2] 
    
    print(lm.beta(mod))
    
    # also do a correlation test - used to visualize 
    # (note - correlation coefficients are the same as standard effect sizes, so this wasn't really needed)
    cor <- cor.test(dat[,pred], dat[,trait])
    res.lr$cors[pred, trait] <- cor$estimate
    
    if(res.lr$pvals[pred, trait] <= 0.05){
      res.lr$eff.sig[pred, trait] <- mod.info$coefficients[pred, 'Estimate']
      res.lr$cors.sig[pred, trait] <- cor$estimate
      res.lr$eff.std.sig[pred,trait] <- lm.beta(mod)$standardized.coefficients[2]
    }
    
  }
}

hist(unlist(res.lr$eff), breaks = 20, main = 'Effect Sizes')
hist(unlist(res.lr$eff.std), breaks = 20, main = 'Standardized Effect Sizes')
hist(unlist(res.lr$cors), breaks = 20, main = 'correlations') # same as std effect sizes
hist(unlist(res.lr$cors.sig), main= 'signficant correlations')

# adjust p-values

padj <- p.adjust(unlist(res.lr$pvals), method = 'BH')
res.lr$padj <- as.data.frame(matrix(padj, ncol = ncol(res.lr$pvals)))
colnames(res.lr$padj) <- colnames(res.lr$pvals)
rownames(res.lr$padj) <- rownames(res.lr$pvals)

plot(unlist(res.lr$pvals), unlist(res.lr$padj), xlab = 'pvals', ylab = 'adjusted pvals')

sum(res.lr$pvals < 0.05) # 19
sum(res.lr$pvals < 0.01) # 7
sum(res.lr$padj < 0.05) # 2
sum(res.lr$padj < 0.1) # 2
```

## Tables with results

```{r print_tables}

kable(res.lr$eff.std.sig, label = 'Standardized effects for significant relationships between plasticity and climate/genetics')

kable(res.lr$pvals, label = 'Unadjusted p-values for relationships between plasticity and climate/genetics' )

kable(res.lr$padj, label = 'Adjusted p-values for relationships between plasticity and climate/genetics' )

```

## Plots of variables with that significantly predict trait plasticity

```{r plot_regressions}

# set up clean trait names
trait_names <- colnames(res.lr$eff.std.sig)

trait_names <- gsub('plasticity_rdpi_', '', trait_names)

trait_names <- gsub('Stage2_cGDD', 'Stage2_Bud_Flush_cGDD', trait_names)
trait_names <- gsub('DOY_Stage2', 'Stage2_Bud_Flush_DOY', trait_names)
trait_names <- gsub('DOY_Stage3', 'Stage3_Leaf_Emergence_DOY', trait_names)
trait_names <- gsub('DOY_Stage6', 'Stage6_First_Budset_DOY', trait_names)
trait_names <- gsub('DOY_Stage7', 'Stage7_Lammas_Growth_DOY', trait_names)
trait_names <- gsub('stage7_presence', 'Stage7_Lammas_Growth_presence', trait_names)
trait_names <- gsub('DOY_Stage8', 'Stage8_Budset_after_Lammas_DOY', trait_names)
trait_names <- gsub('DOY_last_budset', 'Stage8_Final_Budset_DOY', trait_names)
trait_names <- gsub('growing_season_days', 'Growing_Season_Days', trait_names)

trait_names <- gsub('licor_', '', trait_names)
trait_names <- gsub('Fm.', "Fm'", trait_names)
trait_names <- gsub('upper', "Adaxial", trait_names)
trait_names <- gsub('lower', "Abaxial", trait_names)
trait_names <- gsub('stomata_ratio', "Stomata_ratio", trait_names)
trait_names <- gsub('leaf', 'Leaf', trait_names)
trait_names <- gsub('Phi', 'Φ', trait_names)

# remove units
trait_names <- gsub('_cm2', '', trait_names)
trait_names <- gsub('_cm', '', trait_names)
trait_names <- gsub('_mm2', '', trait_names)
trait_names <- gsub('_m2', '', trait_names)
trait_names <- gsub('_mm', '', trait_names)
trait_names <- gsub('_um', '', trait_names)
trait_names <- gsub('_g', '', trait_names)

trait_names <- gsub('_', ' ', trait_names)

# check
cbind(colnames(res.lr$eff.std.sig), trait_names)
# make df to join them
trait_names_df <- cbind.data.frame(colnames(res.lr$eff.std.sig), trait_names)
rownames(trait_names_df) <- trait_names_df[,1]
######################

# plot significant relationships for each predictor (row)

# use matrix with significant effects to select variables
# rename matrix for simplicity
sig <- res.lr$eff.std.sig

# loop through predictors
for(p in 1:nrow(sig)){
  
  xvar <- rownames(sig)[p]
  
  # get significant values
  yvars <- colnames(sig)[! is.na(sig[p,])]
  
  # loop through traits
  plots <- list()
  for(t in 1:length(yvars)){
    
    yvar <- yvars[t]
    
    # get correlation and pval
    cortext <- round(sig[xvar, yvar], 2)
    ptext <- round(res.lr$pvals[xvar, yvar], 3)
    if(ptext == '0'){
      ptext <- '<0.001'
    } else{
      ptext <- paste0('= ', ptext)
    }
    
    
    # make main title pretty
    maintitle <- trait_names_df[yvar, 2]
    # wrap long variables
    maintitle <- paste(strwrap(maintitle, width = 22), collapse = '\n')
    
    # make subtitle
    subtitle <- paste('R = ', cortext, '\n', 'p ', ptext, sep = '')
    
    # make xlabel pretty
    xtitle <- paste(toupper(substr(xvar, 1, 1)), substr(xvar, 2, nchar(xvar)), sep="")
    xtitle <- gsub('_', ' ', xtitle)
    
    p1 <- ggplot(data = dat, aes_string(x = xvar, y = yvar, color = 'k2_tricho'))+
      geom_point(size = 5, na.rm = TRUE) +
      scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = F) +
      geom_smooth(method = lm, na.rm = TRUE, col = 'black')+
      ggtitle(yvar) +
      ylab('Plasticity (RDPI)') +
      xlab(xtitle) +
      ggtitle(maintitle, sub = subtitle) #+
      # theme(plot.title = element_text(size = 16),
      #       plot.subtitle = element_text(size = 14))
  
    plots[[t]] <- p1
    names(plots)[t] <- yvar
    
  }
  
  # save set of plots with predictor name
  assign(paste0('plots.', xvar), plots)
  
  # add legend to plot panel so we don't need a legend for each individual plot
  # do this by extracting the legend from a simplified plot using cowplot
  # then add to the end of the list of plots
  leg_plot <- ggplot(data = dat, aes_string(x = xvar, y = yvar, color = 'k2_tricho'))+
    geom_point()+
      scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry')
  
  legend <- cowplot::get_legend(leg_plot)
  plots[[t+1]] <- legend
  p2 <- grid.arrange(grobs = plots, ncol = 3)
  #plot(p2)
  
}

# PC1 plots - for main text

leg_plot <- 
  ggplot(data = dat, aes_string(x = xvar, y = yvar, color = 'k2_tricho'))+
  geom_point(na.rm = TRUE)+
  scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = expression(paste('% ', italic("P. trichocarpa"))))

legend <- cowplot::get_legend(leg_plot)
plots.genetic_PC1[[length(plots.genetic_PC1)+1]] <- legend
p2 <- grid.arrange(grobs = plots.genetic_PC1, ncol = 3)
plot(p2)
#ggsave(p2, file = 'results/plasticity/plasticityRDPI_by_geneticPC1_significant_bigText.png', height = 8, width = 12)


# all other plots together for supplement
p3 <- grid.arrange(grobs = c(plots.genetic_PC2, plots.genetic_PC3, plots.provenance_MAP, plots.provenance_MAT, plots.provenance_TD, legend$grobs[1]), ncol = 4)
plot(p3)
#ggsave(p3, file = 'results/plasticity/plasticityRDPI_by_allExceptAncestry_significant_bigText.png', height = 16, width = 16)

```

## Heatmap of home-climate relationships

```{r heatmap}

# heatmap params
text_cex <- 1
rot <- 90 # angle for column labels

leg_title <- 14 # font size for legend title
leg_label <- 12 # font size for legend labels

########################################################

# Heatmap of correlations
# colors for heatmap
colf <- colorRampPalette(c('red', 'white', 'blue'))

colnames(res.lr$cors.sig) <- gsub('plasticity_rdpi_', '', colnames(res.lr$cors.sig))
colnames(res.lr$padj) <- gsub('plasticity_rdpi_', '', colnames(res.lr$padj))

# trait categories

# row annotation - trait categories
cats <- c(rep('Phenology', 14),  rep('Leaf Morphology', 4), rep('Stomata', 7), rep('Photosynthesis', 5))
cbind(colnames(res.lr$cors.sig), cats)

ha_row <- rowAnnotation(Trait_Category = cats,
                        show_annotation_name = F,
                            col = list(Trait_Category = c('Phenology' = "#e2e260", 'Photosynthesis' = "#2d4030", 'Leaf Morphology' = "#516823", 'Stomata' =  "#88a2b9")))

# cell text showing significance
# cell_text <- res.lr$pvals
# cell_text[res.lr$pvals > 0.05] <- ''
# cell_text[res.lr$pvals <= 0.05 & res.lr$pvals > 0.01] <- '*'
# cell_text[res.lr$pvals <= 0.01 & res.lr$pvals > 0.001] <- '**'
# cell_text[res.lr$pvals<= 0.001] <- '***'
# cell_text <- t(cell_text)

# or just run for adjusted pvals

cell_text <- res.lr$padj
cell_text[res.lr$padj > 0.05] <- ''
cell_text[res.lr$padj <= 0.05 & res.lr$padj > 0.01] <- '*'
cell_text[res.lr$padj <= 0.01 & res.lr$padj > 0.001] <- '**'
cell_text[res.lr$padj<= 0.001] <- '***'
cell_text <- t(cell_text)

############################
# heatmap by standardized effect size

# color scale based on all effect sizes
max_col <- max(unlist(res.lr$eff.std.sig), na.rm = T)
min_col <- min(unlist(res.lr$eff.std.sig), na.rm = T)
colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))


hm1 <- Heatmap(as.matrix(t(res.lr$eff.std.sig)),
        cluster_rows = F,
        cluster_columns = F,
        col = colf,
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = T,
        column_title = 'Predictors',
        row_title = 'Trait plasticity',
        left_annotation = ha_row,
        heatmap_legend_param = list(title = 'Std. Effect Size'),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))})

ComplexHeatmap::draw(hm1, padding = unit(c(2, 30, 2, 15), "mm"))

```

# Relationship between plasticity and home climate, adjusted for species ancestry


```{r correlations_ancestry_adjusted}


##################################
# models estimating climate effects while controlling for ancestry

# possible predictors of plasticity
preds <- c('provenance_MAT', 'provenance_MAP', 'provenance_TD')

# set up dataframes for saving info within a list
res.lr.pt <- list() # results for linear regressions
# setup template dataframe
df <- as.data.frame(matrix(ncol = length(traits), nrow = length(preds)))
colnames(df) <- traits
rownames(df) <- preds

# add duplicate dataframes to list for saving effect size and pvalues
res.lr.pt$eff <- df
res.lr.pt$eff.sig <- df
res.lr.pt$eff.std <- df
res.lr.pt$eff.std.sig <- df
res.lr.pt$pvals <- df
res.lr.pt$rsq <- df
res.lr.pt$cors <- df
res.lr.pt$cors.sig <- df

# loop through each trait/predictor combination, run a regression, and extract info from model


for(t in 1:length(traits)){
  
  trait <- traits[t]
  
  for(p in 1:length(preds)){
    
    pred <- preds[p]
    
    # model
    form <- paste0(trait, '~', pred, '+ k2_tricho')
    mod <- lm(form, data = dat)
    mod.info <- summary(mod)
    print(paste('Model:', trait, '~', pred))
    print(mod.info)
    
    # save model info
    res.lr.pt$rsq[pred, trait] <- mod.info$adj.r.squared
    res.lr.pt$pvals[pred, trait] <- mod.info$coefficients[pred, 'Pr(>|t|)']
    res.lr.pt$eff[pred, trait] <- mod.info$coefficients[pred, 'Estimate']
    
    # standardize coefficients
    # coef for predictor is second object after intercept
    res.lr.pt$eff.std[pred,trait] <- lm.beta(mod)$standardized.coefficients[2] 
    print(lm.beta(mod))
    
    # also do a correlation test - used to visualize 
    cor <- cor.test(dat[,pred], dat[,trait])
    res.lr.pt$cors[pred, trait] <- cor$estimate
    
    if(res.lr.pt$pvals[pred, trait] <= 0.05){
      res.lr.pt$eff.sig[pred, trait] <- mod.info$coefficients[pred, 'Estimate']
      res.lr.pt$cors.sig[pred, trait] <- cor$estimate
      res.lr.pt$eff.std.sig[pred,trait] <- lm.beta(mod)$standardized.coefficients[2]
    }
  }
}


# adjust p-values

padj <- p.adjust(unlist(res.lr.pt$pvals), method = 'BH')
res.lr.pt$padj <- as.data.frame(matrix(padj, ncol = ncol(res.lr.pt$pvals)))
colnames(res.lr.pt$padj) <- colnames(res.lr.pt$pvals)
rownames(res.lr.pt$padj) <- rownames(res.lr.pt$pvals)

plot(unlist(res.lr.pt$pvals), unlist(res.lr.pt$padj))

sum(res.lr.pt$pvals < 0.05) # 12
sum(res.lr.pt$pvals < 0.01) # 1
sum(res.lr.pt$padj < 0.05) # 1
sum(res.lr.pt$padj < 0.1) # 1


###################
# heatmap by standardized effect size

colnames(res.lr.pt$cors.sig) <- gsub('plasticity_rdpi_', '', colnames(res.lr.pt$cors.sig))
colnames(res.lr.pt$padj) <- gsub('plasticity_rdpi_', '', colnames(res.lr.pt$padj))


# row annotation - trait categories
cats <- c(rep('Phenology', 14),  rep('Leaf Morphology', 4), rep('Stomata', 7), rep('Photosynthesis', 5))
cbind(colnames(res.lr.pt$cors.sig), cats)

ha_row <- rowAnnotation(Trait_Category = cats,
                        show_annotation_name = F,
                            col = list(Trait_Category = c('Phenology' = "#e2e260", 'Photosynthesis' = "#2d4030", 'Leaf Morphology' = "#516823", 'Stomata' =  "#88a2b9")))

# cell text showing significance
# cell_text <- res.lr.pt$pvals
# cell_text[res.lr.pt$pvals > 0.05] <- ''
# cell_text[res.lr.pt$pvals <= 0.05 & res.lr.pt$pvals > 0.01] <- '*'
# cell_text[res.lr.pt$pvals <= 0.01 & res.lr.pt$pvals > 0.001] <- '**'
# cell_text[res.lr.pt$pvals<= 0.001] <- '***'
# cell_text <- t(cell_text)

# or just run for adjusted pvals

cell_text <- res.lr.pt$padj
cell_text[res.lr.pt$padj > 0.05] <- ''
cell_text[res.lr.pt$padj <= 0.05 & res.lr.pt$padj > 0.01] <- '*'
cell_text[res.lr.pt$padj <= 0.01 & res.lr.pt$padj > 0.001] <- '**'
cell_text[res.lr.pt$padj<= 0.001] <- '***'
cell_text <- t(cell_text)

# heatmap by standardized effect size

# color scale based on all effect sizes
max_col <- max(unlist(res.lr.pt$eff.std.sig), na.rm = T)
min_col <- min(unlist(res.lr.pt$eff.std.sig), na.rm = T)
colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))


hm2 <- Heatmap(as.matrix(t(res.lr.pt$eff.std.sig)),
        cluster_rows = F,
        cluster_columns = F,
        #col = colf(100),
        col = colf,
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = T,
        column_title = 'Predictors',
        row_title = 'Trait plasticity',
        left_annotation = ha_row,
        heatmap_legend_param = list(title = 'Std. Effect Size', at = c(-0.6, 0, 0.6)),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))})

ComplexHeatmap::draw(hm2, padding = unit(c(2, 30, 2, 15), "mm"))


```

## Tables with results

```{r print_tables_ancestry_adjusted}

kable(res.lr.pt$eff.std.sig, label = 'Standardized effects for significant relationships between plasticity and climate, controlling for ancestry')

kable(res.lr$pvals, label = 'Unadjusted p-values for relationships between plasticity and climate, controlling for ancestry' )

kable(res.lr$padj, label = 'Adjusted p-values for relationships between plasticity and climate, controlling for ancestry' )


```

# Heatmaps summarizing significant relationships

## Plot plasticity heatmaps together

```{r plot_together}

########## plot together

# rename traits

trait_names <- colnames(res.lr$eff.std.sig)
trait_names <- gsub('plasticity_rdpi_', '', trait_names)
trait_names <- gsub('Stage2_cGDD', 'Stage2_Bud_Flush_cGDD', trait_names)
trait_names <- gsub('DOY_Stage2', 'Stage2_Bud_Flush_DOY', trait_names)
trait_names <- gsub('DOY_Stage3', 'Stage3_Leaf_Emergence_DOY', trait_names)
trait_names <- gsub('DOY_Stage6', 'Stage6_First_Budset_DOY', trait_names)
trait_names <- gsub('DOY_Stage7', 'Stage7_Lammas_Growth_DOY', trait_names)
trait_names <- gsub('stage7_presence', 'Stage7_Lammas_Growth_presence', trait_names)
trait_names <- gsub('DOY_Stage8', 'Stage8_Budset_after_Lammas_DOY', trait_names)
trait_names <- gsub('DOY_last_budset', 'Stage8_Final_Budset_DOY', trait_names)
trait_names <- gsub('growing_season_days', 'Growing_Season_Days', trait_names)
trait_names <- gsub('_', ' ', trait_names)

effect_names1 <- rownames(res.lr$eff.std.sig)
effect_names1 <- gsub('genetic', 'Genetic', effect_names1)
effect_names1 <- gsub('provenance', 'Home', effect_names1)
effect_names1 <- gsub('_', ' ', effect_names1)




########

#draw(hm2 + hm4, padding = unit(c(2, 30, 2, 15), "mm"))

# use same color scale for both heatmaps

# color scale based on all effect sizes
max_col <- max(c(unlist(res.lr$eff.std.sig), unlist(res.lr.pt$eff.std.sig)), na.rm = T)
min_col <- min(c(unlist(res.lr$eff.std.sig), unlist(res.lr.pt$eff.std.sig)), na.rm = T)

# make scale same at both ends
max_col <- max(abs(max_col), abs(min_col))
min_col <- min(abs(max_col), abs(min_col))*-1
max_ceil <- ceiling(max_col*10)/10 # take ceiling rounded to one decimal place - used for placing ticks on legend

colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))


# standardized effect sizes for simple linear regressions

colnames(res.lr$eff.std.sig) <- gsub('plasticity_rdpi_', '', colnames(res.lr$eff.std.sig))

# row annotation - trait categories
cats <- c(rep('Phenology', 14),  rep('Leaf Morphology', 4), rep('Stomata', 7), rep('Photosynthesis', 5))
cbind(colnames(res.lr$cors.sig), cats)

ha_row <- rowAnnotation(Trait_Category = cats,
                        show_annotation_name = F,
                            col = list(Trait_Category = c('Phenology' = "#e2e260", 'Photosynthesis' = "#2d4030", 'Leaf Morphology' = "#516823", 'Stomata' =  "#88a2b9")))

# cell text showing significance
cell_text <- res.lr$padj
cell_text[res.lr$padj > 0.05] <- ''
cell_text[res.lr$padj <= 0.05 & res.lr$padj > 0.01] <- '*'
cell_text[res.lr$padj <= 0.01 & res.lr$padj > 0.001] <- '**'
cell_text[res.lr$padj<= 0.001] <- '***'
cell_text <- t(cell_text)



hm.lr <- Heatmap(as.matrix(t(res.lr$eff.std.sig)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = T,
        column_title = 'Plasticity Predictors',
        column_labels = effect_names1,
        column_title_gp = gpar(fontface = 'bold'),
        row_title_gp = gpar(fontface = 'bold'),
        row_title = 'Trait plasticity',
        row_labels = trait_names,
        left_annotation = ha_row,
        row_names_max_width = unit(8, 'cm'),
        heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))})

ComplexHeatmap::draw(hm.lr, padding = unit(c(2, 30, 2, 15), "mm"))


# linear regression, climate controlling for ancestry

effect_names2 <- rownames(res.lr.pt$eff.std.sig)
effect_names2 <- gsub('provenance', 'Home', effect_names2)
effect_names2 <- gsub('_', ' ', effect_names2)

# cell text showing significance
cell_text <- res.lr.pt$padj
cell_text[res.lr.pt$padj > 0.05] <- ''
cell_text[res.lr.pt$padj <= 0.05 & res.lr.pt$padj > 0.01] <- '*'
cell_text[res.lr.pt$padj <= 0.01 & res.lr.pt$padj > 0.001] <- '**'
cell_text[res.lr.pt$padj<= 0.001] <- '***'
cell_text <- t(cell_text)


hm.lr.pt <- Heatmap(as.matrix(t(res.lr.pt$eff.std.sig)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = 90,
        column_names_centered = T,
        column_title = paste0('Plasticity Predictors', '\n', '(controlling for ancestry)'),
        column_labels = effect_names2,
        column_title_gp = gpar(fontface = 'bold'),
        row_title_gp = gpar(fontface = 'bold'),
        row_title = 'Trait plasticity',
        row_labels = trait_names,
        left_annotation = ha_row,
        #heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        show_heatmap_legend = F, # legend is the same, so don't plot it
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = 1.5))})

ComplexHeatmap::draw(hm.lr.pt, padding = unit(c(2, 30, 2, 15), "mm"))


#png(file = 'results/plasticity/heatmap_plasticity_vs_ancestry_and_climate_padjusted.png', height = 8, width = 10, res = 300, units = 'in')
draw(hm.lr + hm.lr.pt, padding = unit(c(2, 15, 2, 15), "mm"))
#dev.off()




```

## Merge with GxE heatmap

```{r heatmap_with_GxE}

# gxe_hm
# GxE plot has cGDD but not presence/absences traits (stage7 and upper stomata)
load('results/plasticity/GxE_traits_heatmap_data.Rdata')


str(gxe_hm$effect)
str(res.lr$eff.std.sig)

# effect size dataframes
df1 <- gxe_hm$effect
df2 <- res.lr$eff.std.sig
df3 <- res.lr.pt$eff.std.sig

dim(df1)
dim(df2)
dim(df3)

# pvalue dataframes
pval1 <- gxe_hm$pval_adj
pval2 <- res.lr$padj
pval3 <- res.lr.pt$padj

# whitch traits don't match between the two datasets?
colnames(df2)[! colnames(df2) %in% colnames(df1)]
colnames(df1)[! colnames(df1) %in% colnames(df2)]

# remove plasticity from column names
colnames(df2) <- gsub('plasticity_rdpi_', '', colnames(df2))
colnames(df3) <- gsub('plasticity_rdpi_', '', colnames(df3))
colnames(pval2) <- gsub('plasticity_rdpi_', '', colnames(pval2))
colnames(pval3) <- gsub('plasticity_rdpi_', '', colnames(pval3))

# insert empty column for stage 7 presence in GxE heatmap
df1$stage7_presence_2023 <- NA
df1$stage7_presence_2024 <- NA
df1$upper_stomata_presence  <- NA

pval1$stage7_presence_2023 <- NA
pval1$stage7_presence_2024 <- NA
pval1$upper_stomata_presence  <- NA

# insert empty column for GDD for plasticity heatmap

df2$Stage2_cGDD_2023 <- NA
df2$Stage2_cGDD_2024 <- NA
df2$Stage3_cGDD_2023 <- NA
df2$Stage3_cGDD_2024 <- NA

df3$Stage2_cGDD_2023 <- NA
df3$Stage2_cGDD_2024 <- NA
df3$Stage3_cGDD_2023 <- NA
df3$Stage3_cGDD_2024 <- NA

# pvals
pval2$Stage2_cGDD_2023 <- NA
pval2$Stage2_cGDD_2024 <- NA
pval2$Stage3_cGDD_2023 <- NA
pval2$Stage3_cGDD_2024 <- NA

pval3$Stage2_cGDD_2023 <- NA
pval3$Stage2_cGDD_2024 <- NA
pval3$Stage3_cGDD_2023 <- NA
pval3$Stage3_cGDD_2024 <- NA

cbind(colnames(df1), colnames(df2))

# reorder

dput(colnames(df1))

ord1 <- c("growing_season_days_2023", "growing_season_days_2024", "DOY_Stage2_2023", 
"Stage2_cGDD_2023", "DOY_Stage2_2024", "Stage2_cGDD_2024", "DOY_Stage3_2023", 
"Stage3_cGDD_2023", "DOY_Stage3_2024", "Stage3_cGDD_2024", "DOY_Stage6_2023", 
"DOY_Stage6_2024", "DOY_Stage7_2023", "DOY_Stage7_2024", "stage7_presence_2023", 
"stage7_presence_2024", "DOY_last_budset_2023", 
"DOY_last_budset_2024", "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", 
"leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", 
"upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", 
"upper_stomata_density_mm2", "upper_stomata_presence", "stomata_ratio", "licor_gsw", "licor_ETR", "licor_Fm.", 
"licor_Fs", "licor_gbw", "licor_PhiPS2")


dput(colnames(df2))
ord2 <- c("growing_season_days_2023", "growing_season_days_2024", "DOY_Stage2_2023", "Stage2_cGDD_2023",
"DOY_Stage2_2024",  "Stage2_cGDD_2024", "DOY_Stage3_2023","Stage3_cGDD_2023", "DOY_Stage3_2024", "Stage3_cGDD_2024", "DOY_Stage6_2023", 
"DOY_Stage6_2024", "DOY_Stage7_2023", "DOY_Stage7_2024", "stage7_presence_2023", 
"stage7_presence_2024", "DOY_last_budset_2023", "DOY_last_budset_2024", 
"leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023_log", "leaf_mass_g_2023_log",
"LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", 
"lower_stomata_density_mm2", "upper_stomata_density_mm2_log", 
"upper_stomata_presence", "upper_stomata_density_over_total_density_log", "licor_gsw", 
"licor_ETR", "licor_Fm.", "licor_Fs", "licor_gbw", 
"licor_PhiPS2")

# check that they are now in the same order
cbind(ord1, ord2)

# reorder
df1 <- df1[,ord1]
pval1 <- pval1[,ord1]

df2 <- df2[,ord2]
pval2 <- pval2[,ord2]
df3 <- df3[,ord2]
pval3 <- pval3[,ord2]

```


## All heatmaps

```{r all_heatmaps, fig.width = 16, fig.height = 16}

# heatmap params
text_cex <- 1
rot <- 90 # angle for column labels

leg_title <- 14 # font size for legend title
leg_label <- 12 # font size for legend labels

########################################################
# GxE

# rename traits

trait_names <- colnames(df1)
trait_names <- gsub('Stage2_cGDD', 'Stage2_Bud_Flush_cGDD', trait_names)
trait_names <- gsub('DOY_Stage2', 'Stage2_Bud_Flush_DOY', trait_names)
trait_names <- gsub('DOY_Stage3', 'Stage3_Leaf_Emergence_DOY', trait_names)
trait_names <- gsub('DOY_Stage6', 'Stage6_First_Budset_DOY', trait_names)
trait_names <- gsub('DOY_Stage7', 'Stage7_Lammas_Growth_DOY', trait_names)
#trait_names <- gsub('stage7_presence', 'Stage7_Lammas_Growth_presence', trait_names)
trait_names <- gsub('DOY_Stage8', 'Stage8_Budset_after_Lammas_DOY', trait_names)
trait_names <- gsub('DOY_last_budset', 'Stage8_Final_Budset_DOY', trait_names)
trait_names <- gsub('growing_season_days', 'Growing_Season_Days', trait_names)

trait_names <- gsub('licor_', '', trait_names)
trait_names <- gsub('Fm.', "Fm'", trait_names)
trait_names <- gsub('upper', "Adaxial", trait_names)
trait_names <- gsub('lower', "Abaxial", trait_names)
trait_names <- gsub('stomata_ratio', "Stomata_ratio", trait_names)
trait_names <- gsub('leaf', 'Leaf', trait_names)

# remove units
trait_names <- gsub('_cm2', '', trait_names)
trait_names <- gsub('_cm', '', trait_names)
trait_names <- gsub('_mm2', '', trait_names)
trait_names <- gsub('_m2', '', trait_names)
trait_names <- gsub('_mm', '', trait_names)
trait_names <- gsub('_um', '', trait_names)
trait_names <- gsub('_g', '', trait_names)

trait_names <- gsub('_', ' ', trait_names)

# check
cbind(colnames(df1), trait_names)


# rename effects
effect_names <- rownames(df1)
effect_names <- gsub('home', 'Home', effect_names)
effect_names <- gsub('garden', 'Garden', effect_names)
effect_names <- gsub('pc', 'Genetic PC', effect_names)
effect_names <- gsub(':', ' x ', effect_names)
effect_names <- gsub('_', ' ', effect_names)
# check
cbind(rownames(df1), effect_names)

# adjusted pvalues
cell_text <- pval1
cell_text[pval1 > 0.05] <- ''
cell_text[pval1 <= 0.05 & pval1 > 0.01] <- '*'
cell_text[pval1 <= 0.01 & pval1 > 0.001] <- '**'
cell_text[pval1<= 0.001] <- '***'
cell_text <- t(cell_text)

# use same color scale for all heatmaps

# color scale based on all effect sizes
max_col <- max(c(unlist(df1), unlist(df2), unlist(df3)), na.rm = T)
min_col <- min(c(unlist(df1), unlist(df2), unlist(df3)), na.rm = T)

# make scale same at both ends
max_col <- max(abs(max_col), abs(min_col))
min_col <- min(abs(max_col), abs(min_col))*-1
max_ceil <- ceiling(max_col*10)/10 # take ceiling rounded to one decimal place - used for placing ticks on legend

colf <- colorRamp2(c('red', 'white', 'blue'), breaks = c(min_col, 0, max_col))

# column annotation - GxE categories

col_cat <-  c(rep('E', 2), rep('G', 5), rep('GxE', 10))
ha_col <- columnAnnotation(Effect_Category = col_cat,
                        show_annotation_name = F,
                        col = list(Effect_Category = c('E' = "#015b58", 'G' = "#4760b8", 'GxE' = "#380069" )),
                        annotation_legend_param = list(title = 'Effect Category', title_gp = gpar(fontsize = leg_title, fontface = "bold"), labels_gp = gpar(fontsize = leg_label)))


# row annotation - trait categories
cats <- c(rep('Phenology', 18),  rep('Leaf Morphology', 4), rep('Stomata', 7), rep('Photosynthesis', 5))
cbind(trait_names, cats)

ha_row <- rowAnnotation(Trait_Category = cats,
                        show_annotation_name = F,
                            col = list(Trait_Category = c('Phenology' = "#e2e260", 'Photosynthesis' = "#2d4030", 'Leaf Morphology' = "#516823", 'Stomata' =  "#88a2b9")),
                        annotation_legend_param = list(title = 'Trait Category', title_gp = gpar(fontsize = leg_title, fontface = "bold"), labels_gp = gpar(fontsize = leg_label)))


# row annotation - trait categories

hm1 <- Heatmap(as.matrix(t(df1)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        row_labels = trait_names,
        column_names_side = 'top',
        column_names_rot = rot,
        column_names_centered = F,
        column_title = 'Trait Predictors',
        column_title_gp = gpar(fontface = 'bold', fontsize = leg_title),
        column_labels = effect_names,
        row_title_gp = gpar(fontface = 'bold', fontsize = leg_title),
        row_title = 'Trait',
        left_annotation = ha_row,
        top_annotation = ha_col,
        heatmap_legend_param = list(title = 'Std. Effect Size',  at = c(max_ceil*-1, 0, max_ceil), title_gp = gpar(fontsize = leg_title, fontface = "bold"), labels_gp = gpar(fontsize = leg_label)),
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = text_cex))}
        )

ComplexHeatmap::draw(hm1, padding = unit(c(2, 30, 2, 15), "mm"))



########## plot together

effect_names1 <- rownames(df2)
effect_names1 <- gsub('genetic', 'Genetic', effect_names1)
effect_names1 <- gsub('provenance', 'Home', effect_names1)
effect_names1 <- gsub('_', ' ', effect_names1)




########


# standardized effect sizes for simple linear regressions

# cell text showing significance
cell_text <- pval2
cell_text[pval2 > 0.05] <- ''
cell_text[pval2 <= 0.05 & pval2 > 0.01] <- '*'
cell_text[pval2 <= 0.01 & pval2 > 0.001] <- '**'
cell_text[pval2 <= 0.001] <- '***'
cell_text <- t(cell_text)


hm2 <- Heatmap(as.matrix(t(df2)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = rot,
        column_names_centered = F,
        column_title = 'Plasticity\nPredictors',
        column_labels = effect_names1,
        column_title_gp = gpar(fontface = 'bold', fontsize = leg_title),
        row_title_gp = gpar(fontface = 'bold', fontsize = leg_title),
        row_title = 'Trait plasticity',
        row_labels = trait_names,
        left_annotation = ha_row,
        row_names_max_width = unit(8, 'cm'),
        #heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        show_heatmap_legend = F, # legend is the same, so don't plot it
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = text_cex))})

ComplexHeatmap::draw(hm2, padding = unit(c(2, 30, 2, 15), "mm"))
ComplexHeatmap::draw(hm1 + hm2, padding = unit(c(2, 30, 2, 15), "mm"))

# linear regression, climate controlling for ancestry

effect_names2 <- rownames(df3)
effect_names2 <- gsub('provenance', 'Home', effect_names2)
effect_names2 <- gsub('_', ' ', effect_names2)

# cell text showing significance
cell_text <- pval3
cell_text[pval3 > 0.05] <- ''
cell_text[pval3 <= 0.05 & pval3 > 0.01] <- '*'
cell_text[pval3 <= 0.01 & pval3 > 0.001] <- '**'
cell_text[pval3<= 0.001] <- '***'
cell_text <- t(cell_text)


hm3 <- Heatmap(as.matrix(t(df3)),
        cluster_rows = F,
        cluster_columns = F,
        rect_gp = gpar(col = "black", lwd = 1),
        col = colf,
        na_col = 'grey80',
        row_names_side = 'left',
        column_names_side = 'top',
        column_names_rot = rot,
        column_names_centered = F,
        column_title = paste0('Plasticity', '\n', 'Predictors', '\n', '(controlling for ancestry)'),
        column_labels = effect_names2,
        column_title_gp = gpar(fontface = 'bold', fontsize = leg_title),
        row_title_gp = gpar(fontface = 'bold', fontsize = leg_title),
        row_title = 'Trait plasticity',
        row_labels = trait_names,
        left_annotation = ha_row,
        #heatmap_legend_param = list(title = 'Std. Effect Size', at = c(max_ceil*-1, 0, max_ceil)),
        show_heatmap_legend = F, # legend is the same, so don't plot it
        cell_fun = function(j, i, x, y, width, height, fill) { # add text to each grid
                      grid.text(cell_text[i, j], x, y, gp = gpar(cex = text_cex))})

ComplexHeatmap::draw(hm3, padding = unit(c(2, 30, 2, 15), "mm"))


# the combined heatmap!

#png(file = 'results/plasticity/heatmap_ALL_GxE_plasticity_vs_ancestry_and_climate_padjusted.png', height = 10, width = 16, res = 300, units = 'in')
#pdf(file = 'results/plasticity/heatmap_ALL_GxE_plasticity_vs_ancestry_and_climate_padjusted.pdf', height = 10, width = 16)

# alternatively - rotated column names
#png(file = 'results/plasticity/heatmap_ALL_GxE_plasticity_vs_ancestry_and_climate_padjusted_columnNamesRotated.png', height = 10, width = 18, res = 300, units = 'in')
#pdf(file = 'results/plasticity/heatmap_ALL_GxE_plasticity_vs_ancestry_and_climate_padjusted_columnNamesRotated.pdf', height = 10, width = 16)

draw(hm1 + hm2 + hm3, padding = unit(c(2, 15, 2, 15), "mm"))

#dev.off()


```


