---
title: "Test for trait-fitness associations"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
    
---

Test whether traits and their plasticity predict growth in each garden and summarize relationships across garden climates.

# Setup

```{r setup}

library(ggplot2)
library(gridExtra) # arrange plots
library(cowplot) # arrange plots
library(ggrepel) # text labels for gardens
library(ggnewscale) # multiple color scales
library(lme4) # models
library(sjPlot) # model plots
library(car) # Anova

sessionInfo()

# load data
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2024.Rdata')


# set negative heights to NA
dat$GrowthIncrement_2024[dat$GrowthIncrement_2024 < 0] <- NA
dat$GrowthIncrement_2023[dat$GrowthIncrement_2023 < 0] <- NA
dat$GrowthIncrement_2022[dat$GrowthIncrement_2022 < 0] <- NA
dat$GrowthIncrement_2021[dat$GrowthIncrement_2021 < 0] <- NA

# remove NA genotypes and those without genetic info
dat <- dat[! is.na(dat$Genotype),]
dat <- dat[! is.na(dat$k2_tricho),]

# estimate mortality at beginning of spring
# if it was marked alive in fall but did not have a height next spring, mark as dead
dat$Survival_spring_2024 <- NA
dat$Survival_spring_2024[dat$Survival_09_2023 == 1 & is.na(dat$PreFlush_Height_cm_2024)] <- 0
dat$Survival_spring_2024[dat$Survival_09_2023 == 1 & ! is.na(dat$PreFlush_Height_cm_2024)] <- 1

table(dat$Survival_spring_2024, dat$MiniCG_Site)

# load plasticity values - named rdpi
load('data/clean/genotypePlasticityRDPI_2023-2024.Rdata')

# print session info, including package versions
sessionInfo()

```

```{r markdown_settings}

# setup markdown
knitr::opts_chunk$set(fig.width = 14, fig.height = 12)

```

```{r setup_ggplot}

# setup ggplot
theme_set(
  theme_bw(base_size = 12) +
    theme(plot.title = element_text(size = 12))
)

```


```{r setup_plots}

# look at relationships among garden climate variables

# garden climates
par(mfrow = c(1, 3))
plot(dat$garden_climate_2023_PC1, dat$garden_climate_2023_PC2, type = 'n')
text(dat$garden_climate_2023_PC1, dat$garden_climate_2023_PC2, labels = dat$MiniCG_Site)

plot(dat$garden_MAT_2023, dat$garden_MAP_2023, type = 'n')
text(dat$garden_MAT_2023, dat$garden_MAP_2023, labels = dat$MiniCG_Site)

plot(dat$garden_MAT_2023, dat$garden_TD_2023, type = 'n')
text(dat$garden_MAT_2023, dat$garden_TD_2023, labels = dat$MiniCG_Site)
par(mfrow = c(1, 1)) # reset


# plot traits vs fitness for each garden separately

traits <- c("DOY_last_budset_2023", "growing_season_days_2023", "licor_gsw", "licor_gbw", "licor_ETR", "licor_PhiPS2", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023", "leaf_area_cm2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2", "lower_stomata_density_mm2", "stomata_density_both", "upper_stomata_density_over_total_density", "stomata_ratio")

dat$log_GrowthIncrement_2023 <- log(dat$GrowthIncrement_2023+1)

```

## Functions

```{r functions}


# function to convert numeric pvalues to text (used for plot titles)
pval_stars <- function(pval){
  
  if(pval > 0.05){
    return('NS')
  } else if(pval <= 0.05 & pval > 0.01){
    return('*')
  } else if(pval <= 0.01 & pval > 0.001){
    return('**')
  } else if(pval <= 0.001){
    return('***')
  }
  
}


```

## Setup for climate color palettes

```{r setup_colors}

# get climate ranges across both years, so the same color scale is used for 2023 and 2024 plots
# this is used to set 'limits' in scale_color_gradient2() 


# MAT
sites.23 <- c("SWMN", "WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")
sites.24 <- c("SWMN", "WI", "ID", "WYO", "EVERGREEN", "MSU", "MORTON", "PENN", "SU")

sites.mat.23 <- dat$garden_MAT_2023[match(sites.23, dat$MiniCG_Site)]
sites.mat.24 <- dat$garden_MAT_2024[match(sites.24, dat$MiniCG_Site)]

range_mat <- range(c(sites.mat.23, sites.mat.24))

# MAP
sites.map.23 <- dat$garden_MAP_2023[match(sites.23, dat$MiniCG_Site)]
sites.map.24 <- dat$garden_MAP_2024[match(sites.24, dat$MiniCG_Site)]

range_map <- range(c(sites.map.23, sites.map.24))

# TD
sites.td.23 <- dat$garden_TD_2023[match(sites.23, dat$MiniCG_Site)]
sites.td.24 <- dat$garden_TD_2024[match(sites.24, dat$MiniCG_Site)]

range_td <- range(c(sites.td.23, sites.td.24))

```

## Setup trait labels

```{r cleanup_trait_names}

traits <- c("DOY_Stage2_2023", "DOY_Stage3_2023", "DOY_Stage6_2023", "DOY_Stage7_2023", "stage7_presence_2023", "DOY_last_budset_2023", "growing_season_days_2023", "licor_gsw", "licor_gbw", "licor_ETR", "licor_PhiPS2", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023", "leaf_area_cm2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2", "lower_stomata_density_mm2", "stomata_ratio", "DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2024", "DOY_Stage7_2024", "stage7_presence_2024", "DOY_last_budset_2024", "growing_season_days_2024", "plasticity_rdpi_GrowthIncrement_2023_log", "plasticity_rdpi_DOY_Stage2_2023", "plasticity_rdpi_DOY_Stage3_2023", "plasticity_rdpi_DOY_Stage6_2023", "plasticity_rdpi_DOY_last_budset_2023", "plasticity_rdpi_stage7_presence_2023", "plasticity_rdpi_growing_season_days_2023", "plasticity_rdpi_DOY_Stage7_2023", "plasticity_rdpi_licor_gsw", "plasticity_rdpi_licor_gbw", "plasticity_rdpi_licor_PhiPS2", "plasticity_rdpi_licor_ETR", "plasticity_rdpi_licor_Fs", "plasticity_rdpi_licor_Fm.", "plasticity_rdpi_leaf_thickness_avg_mm_2023", "plasticity_rdpi_LMA_g_m2_2023", "plasticity_rdpi_leaf_mass_g_2023_log", "plasticity_rdpi_leaf_area_cm2_2023_log", "plasticity_rdpi_lower_stomata_count", "plasticity_rdpi_lower_stomata_pore_length_mean_um", "plasticity_rdpi_upper_stomata_presence", "plasticity_rdpi_upper_stomata_count_log", "plasticity_rdpi_upper_stomata_pore_length_mean_um", "plasticity_rdpi_upper_stomata_density_mm2_log", "plasticity_rdpi_lower_stomata_density_mm2", "plasticity_rdpi_stomata_density_both", "plasticity_rdpi_upper_stomata_density_over_total_density_log", "plasticity_rdpi_stomata_ratio_log", "plasticity_rdpi_DOY_Stage2_2024", "plasticity_rdpi_DOY_Stage3_2024", 
"plasticity_rdpi_DOY_Stage6_2024", "plasticity_rdpi_DOY_Stage7_2024", 
"plasticity_rdpi_DOY_last_budset_2024", "plasticity_rdpi_stage7_presence_2024", 
"plasticity_rdpi_growing_season_days_2024")

trait_names <- traits
################

trait_names <- gsub('Stage2_cGDD', 'Stage2_Bud_Flush_cGDD', trait_names)
trait_names <- gsub('DOY_Stage2', 'Stage2_Bud_Flush_DOY', trait_names)
trait_names <- gsub('DOY_Stage3', 'Stage3_Leaf_Emergence_DOY', trait_names)
trait_names <- gsub('DOY_Stage6', 'Stage6_First_Budset_DOY', trait_names)
trait_names <- gsub('DOY_Stage7', 'Stage7_Lammas_Growth_DOY', trait_names)
trait_names <- gsub('stage7_presence', 'Stage7_Lammas_Growth_presence', trait_names)
trait_names <- gsub('DOY_Stage8', 'Stage8_Budset_after_Lammas_DOY', trait_names)
trait_names <- gsub('DOY_last_budset', 'Stage8_Final_Budset_DOY', trait_names)
trait_names <- gsub('growing_season_days', 'Growing_Season_Days', trait_names)

trait_names <- gsub('licor_', '', trait_names)
trait_names <- gsub('Fm.', "Fm'", trait_names)
trait_names <- gsub('upper', "Adaxial", trait_names)
trait_names <- gsub('lower', "Abaxial", trait_names)
trait_names <- gsub('stomata_ratio', "Stomata_ratio", trait_names)
trait_names <- gsub('leaf', 'Leaf', trait_names)
trait_names <- gsub('Phi', 'Φ', trait_names)

# remove units
trait_names <- gsub('_cm2', '', trait_names)
trait_names <- gsub('_cm', '', trait_names)
trait_names <- gsub('_mm2', '', trait_names)
trait_names <- gsub('_m2', '', trait_names)
trait_names <- gsub('_mm', '', trait_names)
trait_names <- gsub('_um', '', trait_names)
trait_names <- gsub('_g_', ' ', trait_names)

trait_names <- gsub('plasticity_rdpi_', 'RDPI_', trait_names)

trait_names <- gsub('_', ' ', trait_names)

# check
cbind(traits, trait_names)
# make df to join them
trait_names_df <- cbind.data.frame(traits, trait_names)
rownames(trait_names_df) <- trait_names_df[,1]



```

# Statistical models for traits

Test whether traits predict growth in each garden, and whether the trait-growth relationship varies by garden climate.


## Do trait values predict growth in 2023?

```{r model_growth_increment_2023}

# get 2023 traits

traits <- c("DOY_Stage2_2023", "DOY_Stage3_2023", "DOY_Stage6_2023", "DOY_Stage7_2023", "stage7_presence_2023", "DOY_last_budset_2023", "growing_season_days_2023", "licor_gsw", "licor_gbw", "licor_ETR", "licor_PhiPS2", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023", "leaf_area_cm2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2", "lower_stomata_density_mm2", "stomata_ratio")


# do these traits predict growth increment?

```

### Garden MAT

Test whether garden climate, traits, and their interaction predict growth. An interaction effect suggests that the relationship between the trait and growth varies across garden climates.

This analysis is pretty much repeated for traits, plasticity, each year, and three climate variables (MAT, MAP, TD). If modifying this code in the future - it would probably have been cleaner to create functions instead of repeating loops for each combination.

```{r trait_MAT_2023}

# reorder gardens by garden MAT
tmp <- aggregate(dat$garden_MAT_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

# sites <- levels(dat$MiniCG_Site_2023)
sites <- c("SWMN", "WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")

sites.mat <- dat$garden_MAT_2023[match(sites, dat$MiniCG_Site)]


# list to save results
mods.mat <- list()
# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

# loop through traits
for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))
  
  mod <- lmer(paste0('GrowthIncrement_2023 ~', trait,  '* garden_MAT_2023 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat)
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)

  mods.mat[[n]] <- mod
  names(mods.mat)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  # 1 is trait effect, 2 is garden climate effect, 4 is trait x climate effect.
  # Skip 3 which is ancestry - we already know this predicts height.
  
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    # list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('mat', 'slope', 'pval')
    df.posthoc$mat <- sites.mat
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2023 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')   
        
        # plot data for individual garden
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2023', x = trait, color = 'k2_tricho'), na.rm = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
           'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
           'Trait x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = mat), na.rm = TRUE) +
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAT') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = mat, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    first <- FALSE # if we have made a plot, the next one isn't first

    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = mat, y = slope),  na.rm = TRUE) +
      geom_point(aes(color = mat))+
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}

plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 4)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAT_2023.png', p3, height = 12, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAT_2023.pdf', p3, height = 12, width = 14)


# rename plots list for later use
plots.trait.mat.23 <- plots.summ

```


### Garden MAP

```{r trait_MAP_2023}

###################

# with garden MAP

# reorder gardens by garden MAP
tmp <- aggregate(dat$garden_MAP_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites.map <- dat$garden_MAP_2023[match(sites, dat$MiniCG_Site)]


# list to save results
mods.map <- list()
# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

# loop through traits
for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))

  
  mod <- lmer(paste0('GrowthIncrement_2023 ~', trait,  '* garden_MAP_2023 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat)
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)

  mods.map[[n]] <- mod
  names(mods.map)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  # 1 is trait effect, 2 is garden climate effect, 4 is trait x climate effect.
  # Skip 3 which is ancestry - we already know this predicts height.
  
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    # list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('map', 'slope', 'pval')
    df.posthoc$map <- sites.map
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2023 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')  
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2023', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Trait x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
     # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
            geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = map), na.rm = TRUE) +
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map,  name = 'Garden MAP') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAP') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = map, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
      theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = map))+
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}

# significant interaction for DPY stage6, DOY last budset, gsw, gbw, PhiPS2, Fs, LMA, leaf mass, leaf area, lower pore length, upper pore length
plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 4)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAP_2023.png', p3, height = 12, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAP_2023.pdf', p3, height = 12, width = 14)

# rename plots list for later use
plots.trait.map.23 <- plots.summ

```

### Garden TD

```{r trait_TD_2023}

###################


# with garden TD

# reorder gardens by garden TD
tmp <- aggregate(dat$garden_TD_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites.td <- dat$garden_TD_2023[match(sites, dat$MiniCG_Site)]


# list to save results
mods.td <- list()
# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

# loop through traits
for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))

  
  mod <- lmer(paste0('GrowthIncrement_2023 ~', trait,  '* garden_TD_2023 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat)
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)

  mods.td[[n]] <- mod
  names(mods.td)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  # 1 is trait effect, 2 is garden climate effect, 4 is trait x climate effect.
  # Skip 3 which is ancestry - we already know this predicts height.
  
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    # list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('td', 'slope', 'pval')
    df.posthoc$td <- sites.td
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2023 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')  
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2023', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Trait x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
     # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = td), na.rm = TRUE) +
      scale_color_gradient2(high = "orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits = range_td, name = 'Garden TD (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden TD') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = td, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
      theme(plot.title = element_text(size = 15),
            plot.subtitle = element_text(size = 10),
            legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
     # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = td))+
      scale_color_gradient2(high = "orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits = range_td, name = 'Garden TD (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}

# significant interaction for DPY stage6, DOY last budset, gsw, gbw, PhiPS2, Fs, LMA, leaf mass, leaf area, lower pore length, upper pore length
plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 4)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxTD_2023.png', p3, height = 10, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxTD_2023.pdf', p3, height = 10, width = 14)

# rename plots list for later use
plots.trait.td.23 <- plots.summ

```


## Do trait values predict growth in 2024?


```{r model_growth_increment_2024}

# get 2024 traits
traits <- c("DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2024", "DOY_Stage7_2024", "stage7_presence_2024", "DOY_last_budset_2024", "growing_season_days_2024")


# SWMN was only checked once for stage6 - remove these points
dat[dat$MiniCG_Site == 'SWMN', 'DOY_Stage6_2024'] <- NA
dat[dat$MiniCG_Site == 'SWMN', 'DOY_last_budset_2024'] <- NA

# do these traits predict growth increment?

```


### Garden MAT
```{r trait_MAT_2024}

# reorder gardens by garden MAT
tmp <- aggregate(dat$garden_MAT_2024, by = list(dat$MiniCG_Site_2024), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2024 <- factor(dat$MiniCG_Site_2024, levels = ord)

# sites <- levels(dat$MiniCG_Site_2024)
sites <- c("WI", "ID", "WYO", "EVERGREEN", "MSU", "MORTON", "PENN", "SU")

sites.mat <- dat$garden_MAT_2024[match(sites, dat$MiniCG_Site)]

# list to save results
mods.mat <- list()
# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()


# loop through traits
for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))
  
  
  mod <- lmer(paste0('GrowthIncrement_2024 ~', trait,  '* garden_MAT_2024 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat)
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2024 Growth ~ ', trait, sep = ''))
  plot(p1)

  mods.mat[[n]] <- mod
  names(mods.mat)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  # 1 is trait effect, 2 is garden climate effect, 4 is trait x climate effect.
  # Skip 3 which is ancestry - we already know this predicts height.
  
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    # list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('mat', 'slope', 'pval')
    df.posthoc$mat <- sites.mat
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2024 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2024 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make main title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2024', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
      }
    }
    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
           'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
           'Trait x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    # legend position
    #only plot for first graph
    if(first == TRUE){
      leg_pos <- 'left'
    }else{
      leg_pos <- 'none'
    }
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = mat), na.rm = TRUE) +
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAT') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = mat, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
      theme(plot.title = element_text(size = 15),
            plot.subtitle = element_text(size = 10),
            legend.position = 'none'
      )
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = mat))+
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), name = 'Garden MAT (°C)', limits = range_mat) +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}

# remove empty plots
plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list
plots.summ[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots.summ, ncol = 4)
plot(p3)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAT_2024.png', p3, height = 4, width = 12)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAT_2024.pdf', p3, height = 4, width = 12)

# rename plots list for later use
plots.trait.mat.24 <- plots.summ

```

### Garden MAP

```{r trait_MAP_2024}

###################


# with garden MAP

# reorder gardens by garden MAP
tmp <- aggregate(dat$garden_MAP_2024, by = list(dat$MiniCG_Site_2024), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2024 <- factor(dat$MiniCG_Site_2024, levels = ord)

sites.map <- dat$garden_MAP_2024[match(sites, dat$MiniCG_Site)]


# list to save results
mods.map <- list()
# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

# loop through traits
for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))
  
  
  mod <- lmer(paste0('GrowthIncrement_2024 ~', trait,  '* garden_MAP_2024 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat)
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2024 Growth ~ ', trait, sep = ''))
  plot(p1)

  mods.map[[n]] <- mod
  names(mods.map)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  # 1 is trait effect, 2 is garden climate effect, 4 is trait x climate effect.
  # Skip 3 which is ancestry - we already know this predicts height.
  
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    # list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('map', 'slope', 'pval')
    df.posthoc$map <- sites.map
    
    # make main title pretty
    maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2024 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2024 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2024', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
         ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Trait x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = map), na.rm = TRUE) +
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      ggtitle(maintitle,
              subtitle = sig_text)+
      xlab('Garden MAP') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = map, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
      theme(plot.title = element_text(size = 15),
            plot.subtitle = element_text(size = 10),
            legend.position = 'none')
    plot(p2)
    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = map), na.rm = TRUE)+
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
  }
}


# significant interaction for DOY stage6, DOY last budset, gsw, gbw, PhiPS2, Fs, LMA, leaf mass, leaf area, lower pore length, upper pore length
plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots.summ[[length(plots.summ)+1]] <- legend

# there is only one significant trait

p3 <- grid.arrange(grobs = plots.summ, ncol = 2)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAP_2024.png', p3, height = 4, width = 6)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAP_2024.pdf', p3, height = 4, width = 6)

# rename plots list for later use
plots.trait.map.24 <- plots.summ

```

### Garden TD

```{r trait_TD_2024}

###################

# with garden TD

# reorder gardens by garden TD
tmp <- aggregate(dat$garden_TD_2024, by = list(dat$MiniCG_Site_2024), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2024 <- factor(dat$MiniCG_Site_2024, levels = ord)

sites.td <- dat$garden_TD_2024[match(sites, dat$MiniCG_Site)]


# list to save results
mods.td <- list()
# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

# loop through traits
for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))
  
  mod <- lmer(paste0('GrowthIncrement_2024 ~', trait,  '* garden_TD_2024 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat)
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2024 Growth ~ ', trait, sep = ''))
  plot(p1)

  mods.td[[n]] <- mod
  names(mods.td)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  # 1 is trait effect, 2 is garden climate effect, 4 is trait x climate effect.
  # Skip 3 which is ancestry - we already know this predicts height.
  
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    # list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('td', 'slope', 'pval')
    df.posthoc$td <- sites.td
    
    # make main title pretty
    maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2024 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2024 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2024', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
         ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE)
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Trait x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 30), collapse = '\n')
    
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(color = td), na.rm = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      scale_color_gradient2(high = "orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits = range_td, name = 'Garden TD (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text)+
      xlab('Garden TD') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = td, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
      theme(plot.title = element_text(size = 15),
            plot.subtitle = element_text(size = 10),
            legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    
     # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = td))+
      scale_color_gradient2(high = "orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits = range_td, name = 'Garden TD (°C)')  +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
  }
}


plots.summ <- plots.summ[lengths(plots.summ)>0]
# add legend to beginning of plot list

plots.summ[[length(plots.summ)+1]] <- legend

# there are two significant traits
p3 <- grid.arrange(grobs = plots.summ, ncol = 3)
plot(p3)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxTD_2024.png', p3, height = 4, width = 9)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxTD_2024.pdf', p3, height = 4, width = 9)

# rename plots list for later use
plots.trait.td.24 <- plots.summ

```

## Combine plots for both years

```{r trait_plots_both_years}

# MAT
plots.trait.mat <- c(plots.trait.mat.23, plots.trait.mat.24)
plots.trait.mat <- plots.trait.mat[lengths(plots.trait.mat)>0]

# Because of ggrepel for garden labels, it will not plot if the window is too small
# To fix this when plotting within Rstudio, create a new window and expand to a larger size.
#dev.new()
p.mat <- grid.arrange(grobs = plots.trait.mat, ncol = 4)
plot(p.mat)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAT.png', p.mat, height = 16, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAT.pdf', p.mat, height = 16, width = 14)

# MAP

plots.trait.map <- c(plots.trait.map.23, plots.trait.map.24)
plots.trait.map <- plots.trait.map[lengths(plots.trait.map)>0]

#dev.new()
p.map <- grid.arrange(grobs = plots.trait.map, ncol = 4)
plot(p.map)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAP.png', p.map, height = 14, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxMAP.pdf', p.map, height = 14, width = 14)

# TD

plots.trait.td <- c(plots.trait.td.23, plots.trait.td.24)
plots.trait.td <- plots.trait.td[lengths(plots.trait.td)>0]

#dev.new()
p.td <- grid.arrange(grobs = plots.trait.td, ncol = 4)
plot(p.td)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxTD.png', p.td, height = 14, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitxTD.pdf', p.td, height = 14, width = 14)


```


## Do trait values predict end-of-year survival?


Mostly, no - it seems that there is not enough mortality during one season to predict how trait values affect it

Most of these models did not converge,likely because there was little mortality within the year, so below code is not run


```{r model_survival, eval = F, include = T}

table(dat$Survival_09_2023, dat$MiniCG_Site)

# do these traits predict survival?

# reorder gardens by climate MAT

tmp <- aggregate(dat$garden_MAT_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites <- c("SWMN", "WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")

sites.mat <- dat$garden_MAT_2023[match(sites, dat$MiniCG_Site)]

mods.mat <- list()
for(n in 1:length(traits)){
    
    trait <- traits[n]
    
     print(paste('Model for:', trait))
    
    mod <- glmer(paste0('Survival_09_2023 ~', trait,  '* garden_MAT_2023 + (1 | MiniCG_Site/block) + (1 | Genotype)'),
                data = dat,
                family = binomial
    )
    
    print(car::Anova(mod))
    
    p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std')
    plot(p1)
    #plot_model(mod, type = 're')
    
    mods.mat[n] <- mod
    names(mods.mat)[n] <- trait
    
    # plot data
    p2 <- ggplot(data = dat, aes_string(y = 'Survival_09_2023', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
    geom_point(na.rm = TRUE) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry') +
    ggtitle(trait) +
    stat_smooth(color = 'black', method = "glm", method.args = list(family = "binomial")) +
    facet_wrap(~ MiniCG_Site_2023, drop = T, scales = 'free') +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
    theme(axis.title.x = element_blank())

  plot(p2)
    
}

#plot_models(mods.pc1, vline.color = "black", show.values = T, std.est = T, m.labels = names(mods.pc1))


# with MAP

# reorder gardens by MAP
tmp <- aggregate(dat$garden_MAP_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites <- c("SWMN", "WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")

sites.map <- dat$garden_MAP_2023[match(sites, dat$MiniCG_Site)]


mods.map <- list()

for(n in 1:length(traits)){
    
    trait <- traits[n]
    
    mod <- glmer(paste0('Survival_09_2023 ~', trait,  '* garden_climate_2023_MAP + (1 | MiniCG_Site/block) + (1 | Genotype)'),
                data = dat,
                family = binomial
    )
    
    print(car::Anova(mod))
    
    p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std')
    plot(p1)
    #plot_model(mod, type = 're')
    
      mods.map[n] <- mod
    names(mods.map)[n] <- trait
    
        # plot data
    p2 <- ggplot(data = dat, aes_string(y = 'Survival_09_2023', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
    geom_point(na.rm = TRUE) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry') +
    ggtitle(trait) +
    stat_smooth(color = 'black', method = "glm", method.args = list(family = "binomial")) +
    facet_wrap(~ MiniCG_Site_2023, drop = T, scales = 'free') +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
    theme(axis.title.x = element_blank())

  plot(p2)
    
}

#plot_models(mods.pc2, vline.color = "black", show.values = T, std.est = T, m.labels = names(mods.pc2))




```

# Statistical models for plasticity

Test whether trait plasticity predicts growth in each garden, and whether the plasticity-growth relationship varies by garden climate.

## Does trait plasticity (RDPI) predict growth in 2023?

```{r setup_plasticity_RDPI_2023}

# merge plasticity values with main df

# get plasticity traits
plast <- grep('rdpi', colnames(rdpi), value = T)
# remove 2024 traits
plast <- plast[grep('2024', plast, invert = T)]

# add empty columns
dat[,plast] <- NA

# fill in plasticity columns for each genotype
for(n in 1:nrow(rdpi)){
  
  geno <- rownames(rdpi)[n]
  dat[dat$Genotype == geno, plast] <- rdpi[geno,plast]
  
}


traits <- plast

# get rid of redundant stomata traits 
traits <- traits[! traits %in% c("plasticity_rdpi_lower_stomata_count", "plasticity_rdpi_upper_stomata_count_log", "plasticity_rdpi_upper_stomata_density_over_total_density_log", "plasticity_rdpi_stomata_density_both")]

# does plasticity in these traits predict growth increment?

```

### Garden MAT

```{r plast_MAT_2023}

# reorder gardens by MAT
tmp <- aggregate(dat$garden_MAT_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites <- levels(dat$MiniCG_Site_2023)
# remove UCM - height not measured
sites <- sites[! sites == 'UCM']

sites.mat <- dat$garden_MAT_2023[match(sites, dat$MiniCG_Site)]


# to save results
mods.mat <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  print(paste('Model for:', trait))
  
  mod <- lmer(paste0('GrowthIncrement_2023 ~', trait,  '* garden_MAT_2023 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.mat[[n]] <- mod
  names(mods.mat)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
# list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('mat', 'slope', 'pval')
    df.posthoc$mat <- sites.mat
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2023 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')   
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2023', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Plasticity ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Plasticity x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = mat), na.rm = TRUE) +
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAT') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = mat, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = mat))+
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}


plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 4)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAT_2023.png', p3, height = 12, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAT_2023.pdf', p3, height = 12, width = 18)

# rename plots list for later use
plots.plast.mat.23 <- plots.summ

```

### Garden MAP

```{r plast_MAP_2023}

################################################
# with MAP

# reorder gardens by MAP
tmp <- aggregate(dat$garden_MAP_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites <- levels(dat$MiniCG_Site_2023)
# remove UCM - height not measured
sites <- sites[! sites == 'UCM']

sites.map <- dat$garden_MAP_2023[match(sites, dat$MiniCG_Site)]


# to save results
mods.map <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  mod <- lmer(paste0('GrowthIncrement_2023 ~', trait,  '* garden_MAP_2023 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.map[[n]] <- mod
  names(mods.map)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
# list to save plots for each site
    plots.sub <- list()
    
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('map', 'slope', 'pval')
    df.posthoc$map <- sites.map
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2023 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n') 
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2023', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Plasticity ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Plasticity x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = map), na.rm = TRUE) +
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAP') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = map, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
     # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = map))+
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}



plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 4)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAP_2023.png', p3, height = 6, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAP_2023.pdf', p3, height = 6, width = 14)

# rename plots list for later use
plots.plast.map.23 <- plots.summ

```

### Garden TD

```{r plast_TD_2023}

################################################
# with TD

# reorder gardens by TD
tmp <- aggregate(dat$garden_TD_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites <- levels(dat$MiniCG_Site_2023)
# remove UCM - height not measured
sites <- sites[! sites == 'UCM']

sites.td <- dat$garden_TD_2023[match(sites, dat$MiniCG_Site)]


# to save results
mods.td <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  mod <- lmer(paste0('GrowthIncrement_2023 ~', trait,  '* garden_TD_2023 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.td[[n]] <- mod
  names(mods.td)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
# list to save plots for each site
    plots.sub <- list()
    
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('td', 'slope', 'pval')
    df.posthoc$td <- sites.td
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2023 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n') 
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2023', x = trait, color = 'k2_tricho', na.rm  = TRUE)) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Plasticity ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Plasticity x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = td), na.rm = TRUE) +
      scale_color_gradient2("orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits= range_td, name = 'Garden TD (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden TD') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = td, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = td))+
      scale_color_gradient2("orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits= range_td, name = 'Garden TD (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}


plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 3)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxTD_2023.png', p3, height = 6, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxTD_2023.pdf', p3, height = 6, width = 14)

# rename plots list for later use
plots.plast.td.23 <- plots.summ


```

## Does trait plasticity (RDPI) predict growth in 2024?

```{r setup_plasticity_RDPI_2024}

# merge plasticity values with main df

# get plasticity traits
plast <- grep('rdpi', colnames(rdpi), value = T)
plast <- grep('2024', plast, value = T)

# add empty columns
dat[,plast] <- NA

# fill in plasticity columns for each genotype
for(n in 1:nrow(rdpi)){
  
  geno <- rownames(rdpi)[n]
  dat[dat$Genotype == geno, plast] <- rdpi[geno,plast]
  
}

traits <- plast

# does plasticity in these traits predict growth increment?

```


### Garden MAT

```{r plast_MAT_2024}

# reorder gardens by MAT
tmp <- aggregate(dat$garden_MAT_2024, by = list(dat$MiniCG_Site_2024), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2024 <- factor(dat$MiniCG_Site_2024, levels = ord)

# sites <- levels(dat$MiniCG_Site_2024)
sites <- c("WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")

sites.mat <- dat$garden_MAT_2024[match(sites, dat$MiniCG_Site)]

#sites <- levels(dat$MiniCG_Site_2024)
# remove UCM - height not measured
#sites <- sites[! sites == 'UCM']

# to save results
mods.mat <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  mod <- lmer(paste0('GrowthIncrement_2024 ~', trait,  '* garden_MAT_2024 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2024 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.mat[[n]] <- mod
  names(mods.mat)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
# list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('mat', 'slope', 'pval')
    df.posthoc$mat <- sites.mat
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2024 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        
        mod.sub <- lmer(paste0('GrowthIncrement_2024 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n') 
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2024', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Plasticity ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Plasticity x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = mat), na.rm = TRUE) +
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAT') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = mat, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
      theme(plot.title = element_text(size = 15),
            plot.subtitle = element_text(size = 10),
            legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
     # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = mat))+
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
  }
}


plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list
plots.summ[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots.summ, ncol = 4)
plot(p3)


#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAT_2024.png', p3, height = 8, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAT_2024.pdf', p3, height = 8, width = 14)

# rename plots list for later use
plots.plast.mat.24 <- plots.summ

```

### Garden MAP

```{r plast_MAP_2024}

################################################
# with MAP

# reorder gardens by MAP
tmp <- aggregate(dat$garden_MAP_2024, by = list(dat$MiniCG_Site_2024), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2024 <- factor(dat$MiniCG_Site_2024, levels = ord)

sites <- c("WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")

#sites <- levels(dat$MiniCG_Site_2024)
# remove UCM - height not measured
#sites <- sites[! sites == 'UCM']

sites.map <- dat$garden_MAP_2024[match(sites, dat$MiniCG_Site)]


# to save results
mods.map <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  mod <- lmer(paste0('GrowthIncrement_2024 ~', trait,  '* garden_MAP_2024 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2024 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.map[[n]] <- mod
  names(mods.map)[n] <- trait
  
  #####################
  # posthoc tests
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
# list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('map', 'slope', 'pval')
    df.posthoc$map <- sites.map
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2024 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2024 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')  
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2024', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Plasticity ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Plasticity x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = map), na.rm = TRUE) +
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAP') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = map, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
     # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = map, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = map))+
      scale_color_gradient2(high = "steelblue2",  mid = 'grey', low = "sienna3", midpoint = mean(range(sites.map)), limits = range_map, name = 'Garden MAP (mm)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
  }
}

plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list

plots.summ[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots.summ, ncol = 3)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAP_2024.png', p3, height = 6, width = 10)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAP_2024.pdf', p3, height = 6, width = 10)

# rename plots list for later use
plots.plast.map.24 <- plots.summ

```


### Garden TD

```{r plast_TD_2024}
################################################
# with TD

# reorder gardens by TD
tmp <- aggregate(dat$garden_TD_2024, by = list(dat$MiniCG_Site_2024), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2024 <- factor(dat$MiniCG_Site_2024, levels = ord)

sites <- c("WI", "ID", "WYO", "EVERGREEN", "MSU", "OSU", "MORTON", "WSU", "PENN", "VA", "SU")

#sites <- levels(dat$MiniCG_Site_2024)
# remove UCM - height not measured
#sites <- sites[! sites == 'UCM']

sites.td <- dat$garden_TD_2024[match(sites, dat$MiniCG_Site)]

# to save results
mods.td <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()


for(n in 1:length(traits)){
  
  trait <- traits[n]
  
  mod <- lmer(paste0('GrowthIncrement_2024 ~', trait,  '* garden_TD_2024 + Pt + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2024 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.td[[n]] <- mod
  names(mods.td)[n] <- trait
  
  #####################
  # posthoc tests
  
  # if either the climate or trait x climate are significant, run posthoc tests for each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05 | mod.info$`Pr(>Chisq)`[2] <= 0.05 | mod.info$`Pr(>Chisq)`[4] <=0.05){
    
    
# list to save plots for each site
    plots.sub <- list()
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 3, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('td', 'slope', 'pval')
    df.posthoc$td <- sites.td
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2024 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0('GrowthIncrement_2024 ~', trait, '+ Pt + (1 | block) + (1 | Genotype)'),
                        data = dat.sub)
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect on growth)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # make title pretty
        maintitle <- paste(site, '\n', trait_names_df[trait, 'trait_names'], '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')
        maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')  
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = 'GrowthIncrement_2024', x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(maintitle) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~ x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Plasticity ', pval_stars(mod.info[1, 'Pr(>Chisq)']), '\n',
                       'Climate ', pval_stars(mod.info[2, 'Pr(>Chisq)']), '\n',
                       'Plasticity x Climate ', pval_stars(mod.info[4, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
            geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = td), na.rm = TRUE) +
      scale_color_gradient2("orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits = range_td, name = 'Garden TD (°C)') +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden TD') +
      ylab('Effect on Growth Increment') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = td, y = slope), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none'
)
    
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait

    
     # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = td, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = td))+
      scale_color_gradient2("orchid4",  mid = 'grey', low = "palegreen4", midpoint = mean(range(sites.td)), limits = range_td, name = 'Garden TD (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
  }
}

plots.summ <- plots.summ[lengths(plots.summ)>0]

# add legend to beginning of plot list
plots.summ[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots.summ, ncol = 3)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxTD_2024.png', p3, height = 6, width = 10)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxTD_2024.pdf', p3, height = 6, width = 10)

# rename plots list for later use
plots.plast.td.24 <- plots.summ

```


## Plasticity plots for both years

```{r plasticity_plots_combine_years}

# merge lists of plots created above

plots.plast.mat <- c(plots.plast.mat.23, plots.plast.mat.24)
plots.plast.map <- c(plots.plast.map.23, plots.plast.map.24)
plots.plast.td <- c(plots.plast.td.23, plots.plast.td.24)


# MAT
pMAT <- grid.arrange(grobs = plots.plast.mat, ncol = 4)
plot(pMAT)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAT.png', pMAT, height = 16, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAT.pdf', pMAT, height = 16, width = 14)

# MAP
pMAP <- grid.arrange(grobs = plots.plast.map, ncol = 3)
plot(pMAP)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAP.png', pMAP, height = 12, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxMAP.pdf', pMAP, height = 12, width = 14)

# TD
pTD <- grid.arrange(grobs = plots.plast.td, ncol = 3)
plot(pTD) 
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxTD.png', pTD, height = 12, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityxTD.pdf', pTD, height = 12, width = 14)


```


```{r plot_panel}

# look at overlap in significant traits
g1 <- names(plots.trait.mat)
g2 <- names(plots.plast.mat)
g2 <- gsub('plasticity_rdpi_', '', g2 )
g2 <- gsub('_log', '', g2 )

overlap <- g1[g1 %in% g2]

overlap <- c("DOY_Stage6_2023", "licor_PhiPS2", "licor_Fs", 
"LMA_g_m2_2023", "DOY_last_budset_2024")

# subset of traits with significant interactions

panel <- grid.arrange(grobs = list(plots.trait.mat$leaf_thickness_avg_mm_2023,
                                   plots.trait.mat$DOY_Stage6_2023,
                                   plots.trait.mat$licor_PhiPS2, plots.trait.mat.24[[length(plots.trait.mat.24)]],
                                   plots.plast.mat$plasticity_rdpi_leaf_thickness_avg_mm_2023,
                                   plots.plast.mat.23$plasticity_rdpi_DOY_Stage6_2023 ,                                   plots.plast.mat$plasticity_rdpi_licor_PhiPS2), ncol = 4)

plot(panel)

#ggsave(file = 'results/fitness/fitnessEffectsPanel_MAT_TraitAndPlasticity.png', panel, height = 8, width = 14)
#ggsave(file = 'results/fitness/fitnessEffectsPanel_MAT_TraitAndPlasticity.pdf', panel, height = 8, width = 12)


grid.arrange(grobs = list(plots.trait.mat$LMA_g_m2_2023, plots.plast.mat$plasticity_rdpi_LMA_g_m2_2023), ncol= 2)

grid.arrange(grobs = list(plots.trait.mat$DOY_Stage6_2023, plots.plast.mat$plasticity_rdpi_DOY_Stage6_2023), ncol= 2)

##########################
# MAP
# look at overlap
g1 <- names(plots.trait.map)
g2 <- names(plots.plast.map)
g2 <- gsub('plasticity_rdpi_', '', g2 )
g2 <- gsub('_log', '', g2 )

overlap <- g1[g1 %in% g2]

dput(overlap)

#p <- grid.arrange(grobs = plots.trait.map[overlap], ncol = 4)

panel <- grid.arrange(grobs = list(plots.trait.map$licor_gbw, plots.trait.map$leaf_area_cm2_2023, plots.trait.map$lower_stomata_density_mm2,             plots.plast.map$plasticity_rdpi_licor_gbw, plots.plast.map$plasticity_rdpi_leaf_area_cm2_2023_log, plots.plast.map$plasticity_rdpi_lower_stomata_density_mm2),
                      ncol = 3)
plot(panel)
#ggsave(file = 'results/fitness/fitnessEffectsPanel_MAP_TraitAndPlasticity.png', panel, height = 8, width = 12)

#ggsave(file = 'results/fitness/fitnessEffectsPanel_MAP_TraitAndPlasticity.pdf', panel, height = 8, width = 12)

```



# Is there a tradeoff between plasticity and traits?

If traits and their plasticity have a significant relationship, there could be a tradeoff between plasticity and expression of certain trait values

```{r tradeoff_plasticity_traits}

# list of traits

traits <- c("DOY_Stage2_2023", "DOY_Stage3_2023", "DOY_Stage6_2023", "DOY_Stage7_2023",  "DOY_last_budset_2023", "growing_season_days_2023", "licor_gsw", "licor_gbw", "licor_ETR", "licor_PhiPS2", "licor_Fs", "licor_Fm.", "leaf_thickness_avg_mm_2023", "LMA_g_m2_2023", "leaf_mass_g_2023", "leaf_area_cm2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_presence", "upper_stomata_pore_length_mean_um",  "upper_stomata_density_mm2", "lower_stomata_density_mm2", "stomata_ratio", "DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2024", "DOY_Stage7_2024", "DOY_last_budset_2024", "growing_season_days_2024")

plasts <- paste('plasticity_rdpi_', traits, sep = '')
dput(plasts)

sum(! plasts %in% colnames(dat))
plasts[! plasts %in% colnames(dat)]

# manually edit names of plasticity columns that were log-transformed
dput(plasts)

plasts <- c("plasticity_rdpi_DOY_Stage2_2023", "plasticity_rdpi_DOY_Stage3_2023", 
"plasticity_rdpi_DOY_Stage6_2023", "plasticity_rdpi_DOY_Stage7_2023", 
 "plasticity_rdpi_DOY_last_budset_2023", 
"plasticity_rdpi_growing_season_days_2023", "plasticity_rdpi_licor_gsw", 
"plasticity_rdpi_licor_gbw", "plasticity_rdpi_licor_ETR", "plasticity_rdpi_licor_PhiPS2", 
"plasticity_rdpi_licor_Fs", "plasticity_rdpi_licor_Fm.", "plasticity_rdpi_leaf_thickness_avg_mm_2023", 
"plasticity_rdpi_LMA_g_m2_2023", "plasticity_rdpi_leaf_mass_g_2023_log", 
"plasticity_rdpi_leaf_area_cm2_2023_log", "plasticity_rdpi_lower_stomata_pore_length_mean_um", 
"plasticity_rdpi_upper_stomata_presence", "plasticity_rdpi_upper_stomata_pore_length_mean_um", 
"plasticity_rdpi_upper_stomata_density_mm2_log", "plasticity_rdpi_lower_stomata_density_mm2", 
"plasticity_rdpi_stomata_ratio_log", "plasticity_rdpi_DOY_Stage2_2024", 
"plasticity_rdpi_DOY_Stage3_2024", "plasticity_rdpi_DOY_Stage6_2024", 
"plasticity_rdpi_DOY_Stage7_2024",  
"plasticity_rdpi_DOY_last_budset_2024", "plasticity_rdpi_growing_season_days_2024"
)

cbind(traits, plasts)

################################################
# Do trait values predict trait plasticity, suggesting trade-offs?
# models
################################################
# reorder gardens by MAT
tmp <- aggregate(dat$garden_MAT_2023, by = list(dat$MiniCG_Site_2023), mean)
ord <- tmp$Group.1[order(tmp$x, decreasing = T)]
dat$MiniCG_Site_2023 <- factor(dat$MiniCG_Site_2023, levels = ord)

sites <- levels(dat$MiniCG_Site_2023)
# remove UCM - height not measured
sites <- sites[! sites == 'UCM']

sites.mat <- dat$garden_MAT_2023[match(sites, dat$MiniCG_Site)]


# to save results
mods.mat <- list()

# list to save summary plot for each trait with significant effects on fitness
plots.summ <- list()

for(n in 1:length(traits)){
  
  trait <- traits[n]
  plast <- plasts[n]
  
  print(paste('Model for:', trait, 'and', plast))
  
  # mode;
  mod <- lmer(paste0(trait, '~', plast,  ' + (1 | MiniCG_Site/block) + (1 | Genotype)'),
              data = dat
  )
  
  print(car::Anova(mod))
  
  p1 <- plot_model(mod, vline.color = "black", show.values = T, type = 'std', 
                   title = paste('2023 Growth ~ ', trait, sep = ''))
  plot(p1)
  #plot_model(mod, type = 're')
  
  mods.mat[[n]] <- mod
  names(mods.mat)[n] <- trait
  
  
  #####################
  # posthoc tests
  # if trait significantly predicts plasticity in full model, test in each garden
  mod.info <- Anova(mod)
  # if any coefficients are significant, plot by garden
  # need to change indexes if model is changed!
  if(mod.info$`Pr(>Chisq)`[1] <= 0.05){
    
  # plot data for all sites
  p2 <- ggplot(data = dat, aes_string(y = plast, x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
    geom_point(na.rm = TRUE) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    geom_smooth(color = 'black', lty=1, method = 'lm', formula = y ~ x, na.rm = TRUE) +
    xlab(trait_names_df[trait, "trait_names"]) +
    ylab( paste('Plasticity (RDPI) of', trait_names_df[trait, "trait_names"])) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) #+
  #facet_wrap(~ MiniCG_Site_2023, drop = T, scales = 'free')
  
  plot(p2)
  
   p3 <- ggplot(data = dat, aes_string(y = plast, x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
    geom_point(na.rm = TRUE) +
    scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
    geom_smooth(aes(group = MiniCG_Site_2023), lty=1, method = 'lm', formula = y ~ x, na.rm = TRUE) +
    xlab(trait_names_df[trait, "trait_names"]) +
    ylab( paste('Plasticity (RDPI) of', trait_names_df[trait, "trait_names"])) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12))
    
# list to save plots for each site
    plots.sub <- list()
    
    
    # dataframe to save outputs from posthoc tests
    df.posthoc <- as.data.frame(matrix(ncol = 4, nrow = length(sites)))
    rownames(df.posthoc) <- sites
    colnames(df.posthoc) <- c('mat', 'slope', 'cor', 'pval')
    df.posthoc$mat <- sites.mat
    
    for(s in 1:length(sites)){
      
      site <- sites[s]
      dat.sub <- subset(dat, MiniCG_Site_2023 == site)
      # skip if site is missing data
      if(sum(!is.na(dat.sub[,trait])) > 10){
        mod.sub <- lmer(paste0(trait, '~', plast,  ' + (1 | block) + (1 | Genotype)'),
              data = dat.sub
  )
        # get pval for trait
        mod.sub.info <- Anova(mod.sub)
        df.posthoc[site, 'pval'] <- mod.sub.info$`Pr(>Chisq)`[1]
        
        # get slopes from fixed effect of trait (effect of trait value on plasticity)
        df.posthoc[site, 'slope'] <- fixef(mod.sub)[2]
        
        # get standardized effect, equivalent to correlation
        eff.std <- effectsize::standardize_parameters(mod.sub)
        df.posthoc[site, 'cor'] <- eff.std[2, 'Std_Coefficient'] # 2nd row has coefficient for plasticity effect
        
        # plot
        p <- ggplot(data = dat.sub, aes_string(y = plast, x = trait, color = 'k2_tricho'), na.rm  = TRUE) +
          geom_point(na.rm = TRUE) +
          scale_color_gradient2(high = "darkolivegreen2", mid = "grey20", low = "dodgerblue2", midpoint = 0.5, name = 'P. trichocarpa\nancestry', guide = 'none') +
          ggtitle(paste(site, '\n', trait, '\n', 'p = ', round(df.posthoc$pval[s], 3), sep = '')) +
          geom_smooth(color = 'black', lty = ifelse(df.posthoc$pval[s] <= 0.05, 1,2), method = 'lm', formula = y ~x, na.rm = TRUE) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
          theme(axis.title.x = element_blank())
        
        #plot(p)
        plots.sub[[s]] <- p
        
      }
    }

    # pvals are adjusted for summary plot
    df.posthoc$pvals.adj <- p.adjust(df.posthoc$pval, method = 'BH') 
    
    # plot panel with relationship for each garden
    # remove empty plots
    plots.sub <- plots.sub[lengths(plots.sub)>0]
    # plot in grid
    p1 <- grid.arrange(grobs = plots.sub, ncol = 4)
    #plot(p1)
    
    # setup text showing significance
    sig_text <- paste0('Trait ', pval_stars(mod.info[1, 'Pr(>Chisq)']))
    
    # wrap title
    maintitle <- trait_names_df[trait, 'trait_names']
    maintitle <- paste(strwrap(maintitle, width = 26), collapse = '\n')
    
    # plot slope of fitness relationship vs climate
    p2 <- ggplot(df.posthoc, aes(x = mat, y = cor), na.rm  = TRUE) +
      geom_point(size = 5, stroke = 2, shape = ifelse(df.posthoc$pvals.adj <= 0.05, 16, 1), aes(col = mat), na.rm = TRUE) +
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      geom_smooth(method = 'lm', col = 'black', na.rm = TRUE) +
      geom_hline(yintercept = 0, lty = 2) +
      ggtitle(maintitle,
              subtitle = sig_text) +
      xlab('Garden MAT') +
      ylab('Correlation between Trait and Plasticity') +
      geom_text_repel(label = rownames(df.posthoc), aes(x = mat, y = cor), box.padding = 0.5, min.segment.length = 1, size = 4) +
        theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = 'none')
    plot(p2)
    
    plots.summ[[n]] <- p2
    names(plots.summ)[n] <- trait
    
    # add legend
    leg_plot <- ggplot(df.posthoc, aes(x = mat, y = slope), na.rm  = TRUE) +
      geom_point(aes(color = mat))+
      scale_color_gradient2(high = "red2", mid = "grey", low = "blue2", midpoint = mean(range(sites.mat)), limits = range_mat, name = 'Garden MAT (°C)') +
      theme(legend.position = 'right')
    
    legend <- cowplot::get_legend(leg_plot)
    
    
  }
}


plots.summ <- plots.summ[lengths(plots.summ)>0]

    
# add legend to beginning of plot list

plots2 <- plots.summ
plots2[[length(plots.summ)+1]] <- legend

p3 <- grid.arrange(grobs = plots2, ncol = 4)
plot(p3)

#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityTradeoffMAT2023.png', p3, height = 12, width = 14)
#ggsave(file = 'results/fitness/fitnessEffects_summaryPlot_traitPlasticityTradeoffMAT_2023.pdf', p3, height = 12, width = 14)

# rename plots list for later use
plots.plast.mat.23 <- plots.summ


```

