---
title: "GxE Variance Paritioning"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

Determine the proportion of variance in a trait that is explained by genotype (G), environment (E), genotype x environment (GxE), block nested within garden, and the residuals.

# Setup

```{r setup}

library(ggplot2)
library(lme4)
library(dplyr)
library(tidyr)
library(viridis) # color palettes
library(knitr) # kable() for tables

# load data
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2024.Rdata')

# set negative heights to NA
dat$GrowthIncrement_2023[dat$GrowthIncrement_2024 < 0] <- NA
dat$GrowthIncrement_2023[dat$GrowthIncrement_2023 < 0] <- NA
dat$GrowthIncrement_2022[dat$GrowthIncrement_2022 < 0] <- NA
dat$GrowthIncrement_2021[dat$GrowthIncrement_2021 < 0] <- NA

# remove NA genotypes and those without genetic info
dat <- dat[! is.na(dat$Genotype),]
dat <- dat[! is.na(dat$k2_tricho),]

# print session info, including package versions
sessionInfo()

```


# Statistics - GxE variance partitioning

Get hte proportion of variance explained by factors modelled as random effects using lmer (most traits) or glmer (traits with binary measures).

```{r stats_GxE}

# the list of traits to use

traits <- c("GrowthIncrement_2021", "GrowthIncrement_2022", "GrowthIncrement_2023", "GrowthIncrement_2024", # growth
            "DOY_Stage2_2021", "DOY_Stage3_2021", "DOY_Stage6_2021", "growing_season_days_2021", "Stage2_cGDD_2021", "Stage3_cGDD_2021", # 2021 phenology
            "DOY_Stage2_2022", "DOY_Stage3_2022", "DOY_Stage6_2022", "Stage2_cGDD_2022", "Stage3_cGDD_2022", "growing_season_days_2022", # 2022 phenology
            "DOY_Stage2_2023", "DOY_Stage3_2023", "Stage2_cGDD_2023", "Stage3_cGDD_2023", "DOY_Stage6_2023", "DOY_Stage7_2023", "stage7_presence_2023", "DOY_Stage8_2023", "DOY_last_budset_2023", "growing_season_days_2023", # 2023 phenology
            "DOY_Stage2_2024", "DOY_Stage3_2024", "Stage2_cGDD_2024", 
"Stage3_cGDD_2024", "DOY_Stage6_2024", "DOY_Stage7_2024",  "stage7_presence_2024", "DOY_Stage8_2024", "DOY_last_budset_2024", "growing_season_days_2024",
            "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", "leaf_mass_g_2023", "LMA_g_m2_2023", # leaf morphology
"lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", "upper_stomata_density_mm2", "stomata_ratio", "upper_stomata_presence", "licor_gsw", # stomata 
             "licor_gbw", "licor_ETR", "licor_Fs", "licor_Fm.", "licor_PhiPS2" # photosynthesis
)

# presence/absence traits get modeled differently
traits_bin <- c("stage7_presence_2023", "stage7_presence_2024", "upper_stomata_presence")


# categorize traits for plotting
traits_group <- c(rep('Growth', 4), rep('Phenology', 32), rep('Leaf\nMorphology', 4), rep('Stomata', 7), rep('Photosynthesis', 5))
cbind(traits, traits_group)

# set up dataframes to save results
df <- dat[,traits]
df$genotype <- as.character(dat$Genotype)
df$garden <- as.character(dat$MiniCG_Site)
df$block <- as.character(dat$block)

str(df)

# variance explained df
explained <- as.data.frame(matrix(nrow = length(traits), ncol = 5))
colnames(explained) <- c('Genotype x Environment', 'Genotype', 'Block', 'Environment',  'Residual')
rownames(explained) <- traits


# loop through and fit model to each trait

for(n in 1:length(traits)){
  
  trait <- traits[n]
  print(trait)
  
  # setup formula
  formula <- as.formula(paste(trait, "~ 1 + (1|genotype) + (1|garden/block) + (1|genotype:garden)"))
  
  # presence/absence traits have to be run separately
  # use a glmer with binomial link
  if(trait %in% traits_bin){
    
  # model with a binomial link function
   mod <- glmer(formula,
          data = df,
          family = binomial)
   
    vcs <- as.data.frame(VarCorr(mod))
    
    # calculate proportion variance explained
    # the sum and residuals are (pi^2)/3, see eg:
    # https://stats.stackexchange.com/questions/128750/residual-variance-for-glmer
    vcs_sum <- sum(vcs$vcov) + ((pi^2) / 3)
    explained[n,'Genotype x Environment'] <- (vcs[vcs$grp == 'genotype:garden', 'vcov']/vcs_sum)*100
    explained[n,'Genotype'] <- (vcs[vcs$grp == 'genotype', 'vcov']/vcs_sum)*100
    explained[n,'Environment'] <- (vcs[vcs$grp == 'garden', 'vcov']/vcs_sum)*100
    explained[n, 'Residual'] <- (((pi^2) / 3)/vcs_sum)*100
    explained[n,'Block'] <- (vcs[vcs$grp == 'block:garden', 'vcov']/vcs_sum)*100
    
  }else{
 
  # run the model
  mod <- lmer(formula, data = df)
  
  # save variances
  # need to make sure column names in explained match groups in vcs
  vcs <- as.data.frame(VarCorr(mod))
  explained[n,] <- vcs$vcov
    
  }
}


# convert variance explained to a proportion of the total
explained.prop <- explained

for(n in 1:nrow(explained)){
  
  total <- sum(explained[n, c("Genotype x Environment", "Genotype", "Block", "Environment", "Residual")])
  
  prop <- explained[n, c("Genotype x Environment", "Genotype", "Block", "Environment", "Residual")]/total
  
  explained.prop[n, c("Genotype x Environment", "Genotype", "Block", "Environment", "Residual")] <- prop
  
}

sum(rownames(explained) != rownames(explained.prop))
explained.prop$category <- explained$category

# plot as a stacked barplot

explained$trait <- row.names(explained)
explained$category <- traits_group
exp.long <- explained %>% mutate(trait = factor(trait)) %>% pivot_longer(cols = c('Genotype x Environment', 'Genotype', 'Environment', 'Residual', 'Block'))
exp.long$Effect <- factor(exp.long$name, levels = c('Residual', 'Block', 'Genotype x Environment', 'Genotype', 'Environment'))

ggplot(exp.long, aes(y = trait, x = value, fill = Effect)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_viridis_d(option = 'mako', direction = 1) +
  facet_grid(rows = vars(category), drop = T, scales = 'free_y', space = 'free_y') +
  xlab('Proportion Variance Explained') +
  ylab('Trait') +
  theme_set(theme_bw(base_size = 10))

#ggsave(file = 'results/plasticity/GxE_variance_explained_allTraits_allYears_withBlock.png', height = 10, width = 8)


```

# Table with results

```{r table}

kable(explained, label = 'Raw variance explained')

kable(explained.prop, label = 'Proportion variance explained')

```


# Nice plots

A nicer plot with trait categories and cleaned trait names

```{r nice_plots, fig.height = 8}

# simplify trait names, remove units
colnames(explained)
explained$trait_names <- explained$trait
explained$trait_names <- gsub('Stage2_cGDD', 'Stage2_Bud_Flush_cGDD', explained$trait_names)
explained$trait_names <- gsub('DOY_Stage2', 'Stage2_Bud_Flush_DOY', explained$trait_names)
explained$trait_names <- gsub('DOY_Stage3', 'Stage3_Leaf_Emergence_DOY', explained$trait_names)
explained$trait_names <- gsub('DOY_Stage6', 'Stage6_First_Budset_DOY', explained$trait_names)
explained$trait_names <- gsub('DOY_Stage7', 'Stage7_Lammas_Growth_DOY', explained$trait_names)
explained$trait_names <- gsub('stage7_presence', 'Stage7_Lammas_Growth_presence', explained$trait_names)
explained$trait_names <- gsub('DOY_Stage8', 'Stage8_Budset_after_Lammas_DOY', explained$trait_names)
explained$trait_names <- gsub('DOY_last_budset', 'Stage8_Final_Budset_DOY', explained$trait_names)
explained$trait_names <- gsub('growing_season_days', 'Growing_Season_Days', explained$trait_names)
explained$trait_names <- gsub('licor_', '', explained$trait_names)
explained$trait_names <- gsub('Fm.', "Fm'", explained$trait_names)
explained$trait_names <- gsub('upper', "Adaxial", explained$trait_names)
explained$trait_names <- gsub('lower', "Abaxial", explained$trait_names)
explained$trait_names <- gsub('stomata_ratio', "Stomata_ratio", explained$trait_names)
explained$trait_names <- gsub('leaf', 'Leaf', explained$trait_names)

# remove units
explained$trait_names <- gsub('_cm2', '', explained$trait_names)
explained$trait_names <- gsub('_cm', '', explained$trait_names)
explained$trait_names <- gsub('_mm2', '', explained$trait_names)
explained$trait_names <- gsub('_m2', '', explained$trait_names)
explained$trait_names <- gsub('_mm', '', explained$trait_names)
explained$trait_names <- gsub('_um_', '', explained$trait_names)
explained$trait_names <- gsub('_g', '', explained$trait_names)


explained$trait_names <- gsub('_', ' ', explained$trait_names)

# reorder so they're alphabetical from top to bottom
explained$trait_names <- factor(explained$trait_names, levels = sort(explained$trait_names, decreasing = T))

# transform to long-form dataframe for ggplot
explained$trait <- row.names(explained)
explained$category <- traits_group
exp.long <- explained %>% mutate(trait = factor(trait)) %>% pivot_longer(cols = c('Genotype x Environment', 'Genotype', 'Environment', 'Residual', 'Block'))
exp.long$Effect <- factor(exp.long$name, levels = c('Residual', 'Block', 'Genotype x Environment', 'Genotype', 'Environment'))

# plot!
ggplot(exp.long, aes(y = trait_names, x = value, fill = Effect)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_viridis_d(option = 'mako', direction = 1) +
  facet_grid(rows = vars(category), drop = T, scales = 'free_y', space = 'free_y') +
  xlab('Proportion Variance Explained') +
  ylab('Trait') +
  theme_set(theme_bw(base_size = 10))

#ggsave(file = 'results/plasticity/GxE_variance_explained_allTraits_allYears.png', height = 10, width = 8)


#########################

# subset to only 2023 and 2024 measures
include <- c("GrowthIncrement_2023", "GrowthIncrement_2024","DOY_Stage2_2023", "DOY_Stage3_2023", "Stage2_cGDD_2023", "Stage3_cGDD_2023", "DOY_Stage6_2023", "DOY_Stage7_2023", "stage7_presence_2023", "DOY_Stage8_2023", "DOY_last_budset_2023", "growing_season_days_2023", "DOY_Stage2_2024", "DOY_Stage3_2024", "Stage2_cGDD_2024", "Stage3_cGDD_2024", "DOY_Stage6_2024", "DOY_Stage7_2024", "stage7_presence_2024", "DOY_Stage8_2024", "DOY_last_budset_2024", "growing_season_days_2024", "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", "leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", "upper_stomata_density_mm2", "stomata_ratio", "upper_stomata_presence", "licor_gsw", "licor_gbw", "licor_ETR", "licor_Fs", "licor_Fm.", "licor_PhiPS2")
explained.sub <- explained[explained$trait %in% include,]


# reorder so they're alphabetical from top to bottom
# reversed because geom_bar plots them in reverse
explained.sub$trait_names <- factor(explained.sub$trait_names, levels = sort(explained.sub$trait_names, decreasing = F))

# transform to long-form dataframe for ggplot
exp.sub.long <- explained.sub %>% mutate(trait_names = factor(trait_names)) %>% pivot_longer(cols = c('Genotype x Environment', 'Genotype', 'Environment', 'Residual', 'Block'))
#exp.sub.long$Effect <- factor(exp.sub.long$name, levels =c('Residual', 'Block', 'Genotype x Environment', 'Genotype', 'Environment'))

exp.sub.long$Effect <- factor(exp.sub.long$name, levels =c("Genotype", "Environment", "Genotype x Environment", "Block", "Residual"))

exp.sub.long$category <- factor(exp.sub.long$category, levels = c('Growth', 'Phenology', 'Leaf\nMorphology', 'Stomata', 'Photosynthesis'))

# edit color palette
# I started with this palette: PNWColors::rev(pnw_palette('Starfish', 4))
# then edited colors to be a bit more distinguishable
cols <- c("#4760b8", "#015b58", "#380069", "#e69b99")

ggplot(exp.sub.long, aes(y = trait_names, x = value, fill = Effect)) +
  geom_bar(aes(fill = Effect), position = position_fill(reverse = TRUE), stat = "identity") +
  scale_fill_manual(values = c(cols, 'grey50')) +
  facet_grid(rows = vars(category), drop = T, scales = 'free_y', space = 'free_y') +
  xlab('Proportion Variance Explained') +
  ylab('Trait') +
  theme_set(theme_bw(base_size = 10))


#ggsave(file = 'results/plasticity/GxE_variance_explained_allTraits_2023-2024.png', height = 8, width = 8)

#ggsave(file = 'results/plasticity/GxE_variance_explained_allTraits_2023-2024.pdf', height = 8, width = 8)


##################
# Final plot
# subset to only 2023 and 2024 measures
# remove cGDD

include <- c("GrowthIncrement_2023", "GrowthIncrement_2024","DOY_Stage2_2023", "DOY_Stage3_2023","DOY_Stage6_2023", "DOY_Stage7_2023", "stage7_presence_2023", "DOY_last_budset_2023", "growing_season_days_2023", "DOY_Stage2_2024", "DOY_Stage3_2024", "DOY_Stage6_2024", "DOY_Stage7_2024", "stage7_presence_2024", "DOY_last_budset_2024", "growing_season_days_2024", "leaf_thickness_avg_mm_2023", "leaf_area_cm2_2023", "leaf_mass_g_2023", "LMA_g_m2_2023", "lower_stomata_pore_length_mean_um", "upper_stomata_pore_length_mean_um", "lower_stomata_density_mm2", "upper_stomata_density_mm2", "stomata_ratio", "upper_stomata_presence", "licor_gsw", "licor_gbw", "licor_ETR", "licor_Fs", "licor_Fm.", "licor_PhiPS2")
explained.sub <- explained[explained$trait %in% include,]

# reorder so they're alphabetical from top to bottom
explained.sub$trait_names <- factor(explained.sub$trait_names, levels = sort(explained.sub$trait_names, decreasing = T))

exp.sub.long <- explained.sub %>% mutate(trait_names = factor(trait_names)) %>% pivot_longer(cols = c('Genotype x Environment', 'Genotype', 'Environment', 'Residual', 'Block'))
exp.sub.long$Effect <- factor(exp.sub.long$name, levels =c('Residual', 'Block', 'Genotype x Environment', 'Genotype', 'Environment'))

# edit color palette
# I started with this palette: PNWColors::rev(pnw_palette('Starfish', 4))
# then edited colors to be a bit more distinguishable
cols <- c("#e69b99","#380069","#4760b8","#015b58")

ggplot(exp.sub.long, aes(y = trait_names, x = value, fill = Effect)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = c('grey50', cols)) +
  facet_grid(rows = vars(category), drop = T, scales = 'free_y', space = 'free_y') +
  xlab('Proportion Variance Explained') +
  ylab('Trait') +
  theme_set(theme_bw(base_size = 10))

#ggsave(file = 'results/plasticity/GxE_variance_explained_withBlock_clean_2023-2024.png', height = 8, width = 8)

```


